---
title: "E8.5 seqFISH - Manual reannotation of cell types"
author: "Shila Ghazanfar"
date: "31/07/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 10 
                     fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE, cache.lazy = FALSE)
```

```{r}
set.seed(2020)
```

# Summary

Here I take the cell classification from KNN when integrating with the 10X 
gastrulation atlas, as well as joint subclustering of 10X and seqFISH cells.
I then reassign the KNN based cell type labels based on manual reannotation
from Tim.

There are three important parts:

- reannotate all of joint subclusters

- reassign virtually dissected subset of cluster 6 as Gut tube

- redefine "Def. endoderm" as "Definitive endoderm"

I don't actually need the count data to perform this, just the metadata.

```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
library(ggmap)
library(randomForest)
library(batchelor)
library(igraph)
library(cowplot)
library(ggpubr)
library(data.table)
```

# Load automated cell type annotation

```{r}
knn = readRDS("../analysis_output/E8.5/E8.5_naive_pred_knn_libsize.Rds")
knn[knn == "Def. endoderm"] <- "Definitive endoderm"
celltype_mapped_manual <- knn[,"celltype_mapped"]
```

# Load joint subclustering

```{r}
subclusters_raw = readRDS("../analysis_output/E8.5/E8.5_joint_subclustering_libsize.Rds")
subclusters = subclusters_raw[rownames(knn)]
```

# Reannotate joint subclusters

```{r}
new_annotation = read.delim("../analysis_output/guttubecells/annotation_changes.tsv",
                            header = TRUE, row.names = 1, stringsAsFactors = FALSE)
new_annotation

for (i in 1:nrow(new_annotation)) {
  
  celltype_mapped_manual[subclusters == rownames(new_annotation)[i]] <- new_annotation[i,"New.name"]
  
}
```

# Reassign virtually dissected subset of cluster 6 as Gut tube

```{r}
gut_tube_cells_list = sapply(list.files("../analysis_output/guttubecells/",
                                        pattern = "cells", full.names = TRUE), readRDS, simplify = FALSE)

gut_tube_cells = unlist(gut_tube_cells_list)
length(gut_tube_cells)
length(unique(gut_tube_cells))
table(subclusters[gut_tube_cells])

celltype_mapped_manual[gut_tube_cells] <- "Gut tube"
```

# redefine "Def. endoderm" as "Definitive endoderm"

```{r}
celltype_mapped_manual[celltype_mapped_manual == "Def. endoderm"] <- "Definitive endoderm"
```

# Combine

```{r}
sort(table(celltype_mapped_manual))
table(celltype_mapped_manual == knn[,"celltype_mapped"])

tab = unclass(table(knn[,"celltype_mapped"], celltype_mapped_manual))
tablong = sort_df(melt(tab), "value")
tablong[as.character(tablong[,1]) != as.character(tablong[,2]) & tablong[,3] > 0,]

knn_new <- data.frame(celltype_mapped_refined = factor(celltype_mapped_manual),
                      celltype_mapped = factor(knn[,"celltype_mapped"]),
                      celltype_mapped_alternative = factor(knn[,"celltype_alternative"]),
                      celltype_mapping_score = as.numeric(knn[, "mapping_score"]),
                      celltype_mapping_score_alternative = as.numeric(knn[, "mapping_alternative_score"]))

dim(knn_new)
head(knn_new)
```

# Finish

```{r}
saveRDS(knn_new, file = "../analysis_output/E8.5/celltype_annotation_refined.Rds")
saveRDS(levels(knn_new$celltype_mapped_refined), file = "../analysis_output/E8.5/celltypes_refined.Rds")
```


```{r}
sessionInfo()
```