---
title: "E8.5 smFISH - Generate counts matrix"
author: "Shila Ghazanfar"
date: "29/09/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 10 
                     fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE,cache.lazy = FALSE)
```

```{r}
set.seed(2020)
```

```{r}
library(ggplot2)
library(pracma)
require(rhdf5)
```

```{r}
library(SpatialUtils)
```

Note that here i call objects by "lower" but there is just a single
thresholding selection.

# Directories

```{r}
segmentation_data_dir_embryo12 = "../Data/TimEmbryos-102219/segmentation/"
fov_info_filename_embryo12 = "../Data/TimEmbryos-102219/output/cellinfo/fovinfo.csv"
smFISH_lower_data_dir_embryo12 = "../Data/smFISH/calibrating/thresh_embryo12/"

segmentation_data_dir_embryo3 = "../Data/TimEmbryos-120919/segmentation/"
fov_info_filename_embryo3 = "../Data/TimEmbryos-120919/output/cellinfo/fovinfo.csv"
smFISH_lower_data_dir_embryo3 = "../Data/smFISH/calibrating/thresh_embryo3/"
```

# Gene probe label

```{r}
gene_label_raw = read.delim("../Data/smFISH/channel_info/smFISH__gene_key__new.tab",
                            row.names = 2, header = TRUE)
gene_label = as.character(gene_label_raw$Gene)
names(gene_label) <- rownames(gene_label_raw)
gene_label
```


# Summary

The easiest way to do this would be to grab the polygon segmentation and ask if 
dots fall inside or outside these polygonalisations, but the polygons are 
global transformed, this might actually become really finnicky and maybe even 
incorrect. So instead I work with the dense segmentations.

# Check what coordinate system the mRNA spot calls are in

```{r}
embryofiles = c("embryo12", "embryo3")

pos_valuesList = list("embryo12" = as.character(0:50),
                      "embryo3" = as.character(0:39))

for (embryofile in embryofiles) {

pos_values = pos_valuesList[[embryofile]]

suffix = "_1_1__coords.tab"

coords_list = sapply(pos_values[1:2],
                     function(pos_value) {
                         filename = paste0(get(paste0("smFISH_lower_data_dir_", embryofile)), pos_value, suffix)
                         out = read.csv(filename, header = FALSE)
                         colnames(out) <- c("x","y","z")
                         out$pos = paste0("Pos",pos_value)
                         return(out)
                     }, 
                     simplify = FALSE)
coords = do.call(rbind, coords_list)

ggplot(coords, aes(x = x, y = y, colour = pos)) + 
    geom_point() + 
    theme_classic() + 
    NULL
}
```

These are local coordinates. So we can match these smFISH dots to the 
segmentation directly, no need to perform any affine transformations.

# Function to extract all mRNA spot calls for each field of view

```{r}
gene_values = unique(
    unlist(
        lapply(
            lapply(
                strsplit(list.files(smFISH_lower_data_dir_embryo12, full.names = FALSE, pattern = "coords.tab"),
                                                 "_"), "[", c(2,3)
                ), paste0, collapse = "_")
        )
    )
gene_values
length(gene_values)

extractSpotCalls = function(pos_value, data_dir) {
    
    print(pos_value)
    
    all_calls_list = sapply(gene_values,
                            function(gene_value) {
                                filenames = paste0(data_dir, pos_value, "_",
                                                   gene_value, "_", c("_coords.tab",
                                                                      "_intensity.tab",
                                                                      "_sigma.tab"))
                                
                                if (!file.exists(filenames[1])) return(NULL)
                                
                                out = read.csv(filenames[1], header = FALSE)
                                colnames(out) <- c("x","y","z")
                                out$pos = paste0("Pos", pos_value)
                                out$gene_value <- gene_value
                                
                                intensity = scan(file = filenames[2])
                                out$intensity <- intensity
                                
                                sigma = scan(file = filenames[3])
                                out$sigma <- sigma
                                
                                return(out)
                            }, 
                            simplify = FALSE)
    
    all_calls = do.call(rbind, all_calls_list)
    
    return(all_calls)
}
```


```{r}
all_calls_lower_embryo12 = do.call(
    rbind, sapply(pos_valuesList[["embryo12"]], extractSpotCalls, data_dir = smFISH_lower_data_dir_embryo12, simplify = FALSE))

all_calls_lower_embryo3 = do.call(
    rbind, sapply(pos_valuesList[["embryo3"]], extractSpotCalls, data_dir = smFISH_lower_data_dir_embryo3, simplify = FALSE))
```

# Dense segmentation information 

Take the h5 segmentation masks and assign a unique cell ID to the smFISH dots

```{r}
# manual
embryo_label_12 = c(rep("embryo1", 29), rep("embryo2", 22))
names(embryo_label_12) <- paste0("Pos", 0:50)

embryo_label_3 = rep("embryo3", 40)
names(embryo_label_3) <- paste0("Pos", 0:39)

embryo_labelList = list("embryo12" = embryo_label_12,
                        "embryo3" = embryo_label_3)
```


```{r}
assignToSegmentsPixels = function(embryofile,
                                  spot_calls,
                                  segmentation_data_dir) {
    
    require(rhdf5)
    
    spot_calls$x_pixel = round(spot_calls$x/4)
    spot_calls$y_pixel = round(spot_calls$y/4)
    spot_calls$cellID = NA
    
    pos_values = pos_valuesList[[embryofile]]
    
    for (pos_value in paste0("Pos", pos_values)) {
        
        print(pos_value)
        
        for (z_value in c(2,5)) {
            print(z_value)
            
            h5_file = paste0(segmentation_data_dir, "boundaryseg_", pos_value,
                             "-000",z_value,"_Multicut Segmentation.h5")
            
            file.exists(h5_file)
            
            data_name = h5ls(h5_file)$name
            data = h5read(file = h5_file, data_name)
            
            all_calls_sub_ind = spot_calls$z == z_value & spot_calls$pos == pos_value
            table(all_calls_sub_ind)
            
            x_pixel = spot_calls[all_calls_sub_ind, "x_pixel"]
            y_pixel = spot_calls[all_calls_sub_ind, "y_pixel"]
            
            cellID = sapply(seq_len(length(x_pixel)), function(i) data[x_pixel[i],y_pixel[i],1])
            
            spot_calls[all_calls_sub_ind,"cellID"] <- cellID
            
        }
    }
    
    # after loop
    uniqueID <- paste0(paste0(embryo_labelList[[embryofile]][spot_calls$pos],"_",spot_calls$pos,"_cell",spot_calls$cellID,"_z",spot_calls$z))
    
    spot_calls$uniqueID <- uniqueID
    
    return(spot_calls)
    
}
```

```{r}
all_calls_lower_embryo12 <- assignToSegmentsPixels(
    embryofile = "embryo12", 
    spot_calls = all_calls_lower_embryo12,
    segmentation_data_dir = segmentation_data_dir_embryo12)

all_calls_lower_embryo3 <- assignToSegmentsPixels(
    embryofile = "embryo3", 
    spot_calls = all_calls_lower_embryo3,
    segmentation_data_dir = segmentation_data_dir_embryo3)

all_calls_lower_embryo12 = subset(all_calls_lower_embryo12, !is.na(cellID))
all_calls_lower_embryo3 = subset(all_calls_lower_embryo3, !is.na(cellID))
```

# Generate gene expression matrices of only the smFISH genes

```{r}
all_cells = sort(unique(c(as.character(all_calls_lower_embryo12[,"uniqueID"]),
                          as.character(all_calls_lower_embryo3[,"uniqueID"]))))
length(all_cells)

all_calls_lower = rbind(all_calls_lower_embryo12, all_calls_lower_embryo3)

all_calls_lower$gene_value_name = gene_label[as.character(all_calls_lower$gene_value)]

dim(all_calls_lower)

table(all_calls_lower$gene_value_name)
```

```{r}
buildSMFISHExpressionMatrix = function(spot_calls, all_cells, geneID = "gene_value") {
    # spot_calls is a dataframe with each mRNA spot called
    # all_cells is a character vector of the cells of interest
    # geneID is the column to pull out the gene information
    
    require(reshape)
    
    spot_calls <- subset(spot_calls, uniqueID %in% all_cells)
    spot_calls$uniqueID <- factor(spot_calls$uniqueID, levels = all_cells)
    
    mat = tapply(spot_calls[,1],list(spot_calls[,geneID],
                                     spot_calls$uniqueID),
                 function(x)length(x),
                 default = 0)
    
    return(mat)
}
```

```{r}
smFISH_lower = buildSMFISHExpressionMatrix(all_calls_lower,
                                           all_cells = all_cells,
                                           geneID = "gene_value_name")
dim(smFISH_lower)
smFISH_lower[1:5,1:5]
```

Save the output

```{r}
saveRDS(smFISH_lower, file = "../analysis_output/smFISH_counts.Rds")
```

# Finish

```{r}
sessionInfo()
```
