---
title: "E8.5 seqFISH - Subclustering the brain"
author: "Shila Ghazanfar"
date: "07/09/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 16
                     fig_height: 12
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE, cache.lazy = FALSE)
```

```{r}
set.seed(2020)
```

```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
library(ggmap)
library(randomForest)
library(batchelor)
library(igraph)
library(cowplot)
library(ggpubr)
library(Matrix)
library(grDevices)
library(reshape)
```

```{r}
sapply(list.files("../functions/", full.names = TRUE), source)
```

# Load spatial data

```{r}
sce = readRDS("../analysis_output/E8.5/E8.5_sce_filt_unlabelled.Rds")
sce <- reNormalise(sce)
sce
```

# Add cell type and clustering information

```{r}
mapping_dt = readRDS("../analysis_output/E8.5/celltype_annotation_refined.Rds")
colData(sce) <- cbind(colData(sce), mapping_dt[colnames(sce), setdiff(colnames(mapping_dt),colnames(colData(sce)))])
```

# Load atlas data

```{r}
atlas = readRDS("../analysis_output/E8.5/atlas_for_celltype_mapping.Rds")
atlas
atlas <- scran::computeSumFactors(atlas)
atlas <- scater::normalize(atlas)

# add the UMAP coords to the colData
colData(atlas)$UMAP1 <- reducedDim(atlas, "umap")[,1]
colData(atlas)$UMAP2 <- reducedDim(atlas, "umap")[,2]
```

# Subset the atlas to Forebrain/Midbrain/Hindbrain and cluster further

```{r}
atlas_sub = atlas[, atlas$celltype %in% c("Forebrain/Midbrain/Hindbrain")]

if (!file.exists("../analysis_output/E8.5/atlas_cluster_brain_factor.Rds")) {
  
  assay(atlas_sub, "cosineNorm") <- cosineNorm(assay(atlas_sub, "logcounts"))
  
  stats = modelGeneVar(atlas_sub)
  hvgs = setdiff(getTopHVGs(stats), "Xist")
  print(length(hvgs))
  
  mbpca = multiBatchPCA(assay(atlas_sub, "cosineNorm"),
                        batch = atlas_sub$sample,
                        preserve.single = TRUE,
                        subset.row = hvgs)
  reducedDim(atlas_sub, "mbPCA_cosine") <- mbpca[[1]]
  
  out = reducedMNN(mbpca[[1]], batch = atlas_sub$sample)
  
  reducedDim(atlas_sub, "mbPCA_cosine_MNN") <- out$corrected
  
  snn = buildSNNGraph(atlas_sub, use.dimred = "mbPCA_cosine_MNN")
  clusters_graph = cluster_louvain(snn)
  clusters = as.numeric(membership(clusters_graph))
  names(clusters) <- atlas_sub$cell
  table(clusters, useNA = "always")
  
  atlas_sub$cluster_brain = factor(clusters)
  
  saveRDS(atlas_sub$cluster_brain,
          file = "../analysis_output/E8.5/atlas_cluster_brain_factor.Rds")
} else {
  atlas_sub$cluster_brain <- readRDS("../analysis_output/E8.5/atlas_cluster_brain_factor.Rds")
}
```

```{r}
ggplot(as.data.frame(colData(atlas_sub)),
       aes(x = UMAP1, y = UMAP2, colour = cluster_brain)) + 
  geom_point() + 
  scale_colour_viridis_d() + 
  NULL
```

# Markers for atlas brain subclusters

```{r}
out = getMarkerArray(atlas_sub,
                     group = "cluster_brain",
                     assayName = "logcounts",
                     block = "sample",
                     pseudobulk = TRUE)


out_DE = getMarkerArray(atlas_sub,
                        group = "cluster_brain",
                        assayName = "logcounts",
                        block = "sample",
                        pseudobulk = TRUE,
                        subset = atlas_sub$cluster_brain %in% c(5,6))

markerArrayPlot(out_DE, grouplevel = "5", otherGroupLabel = "6")


for (group in dimnames(out)[[2]]) {
  markerArrayPlot(out,
                  grouplevel = group,
                  otherGroupLabel = "all other cells",
                  LFC = 0.5, FDR = 0.2)
}
```

# Closest cells in seqFISH

Ask for each of the atlas brain cells which spatial cells they most closely 
correspond to

```{r}
closest = readRDS("../analysis_output/E8.5/E8.5_atlas_closest_spatial_cells_libsize.Rds")
closest_cell = readRDS("../analysis_output/E8.5/E8.5_closest_atlas_cell_libsize.Rds")
k.mapped = readRDS("../analysis_output/E8.5/E8.5_naive_k.mapped_libsize.Rds")

k.mapped.brain = vectorMatch(as.character(atlas_sub$cluster_brain),k.mapped,colnames(atlas_sub))
mapped.brain = apply(k.mapped.brain, 1, getmode, dist = 1:ncol(k.mapped.brain))

sce$mapped.brain <- unlist(mapped.brain)[colnames(sce)]
sce$mapped.brain[sce$celltype_mapped_refined != "Forebrain/Midbrain/Hindbrain"] <- NA

table(sce$mapped.brain, useNA = "always")

mapped_brain_df = get_MVS_vals(k.mapped.brain)
sce$mapped.brain.score = mapped_brain_df[, "mapping_score"]
```

A bit under half of the 10X brain cells are the closest to some spatial cells.
Take this subset of spatial cells and plot them in space with the brain cluster.

```{r}
sce_sub = sce[, !is.na(sce$mapped.brain)]
sce_sub

sort(table(sce_sub$celltype_mapped))

table(sce_sub$mapped.brain, useNA = "always")

table(colnames(sce) %in% colnames(sce_sub), sce$celltype_mapped_refined == "Forebrain/Midbrain/Hindbrain")
```

A lot of spatial cells map to the same cell in the atlas, which is not so 
unexpected.

```{r}
ggplot(as.data.frame(colData(sce_sub)),
       aes(x = x_global_affine, 
           y = -y_global_affine)) + 
  geom_point(aes(colour = mapped.brain),
             size = 1.5) + 
  theme_classic() +
  facet_grid(z ~ embryo) + 
  coord_fixed() +
  NULL
```


```{r}
ggplot(as.data.frame(colData(sce)),
       aes(x = x_global_affine, 
           y = -y_global_affine)) + 
  geom_point(aes(colour = mapped.brain),
             size = 1.5, alpha = 0.5) + 
  theme_classic() +
  facet_grid(z ~ embryo) + 
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  coord_fixed() +
  NULL
```

```{r, fig.width=20, fig.height=30}
DF = as.data.frame(colData(sce))
DF <- DF[, setdiff(colnames(df), "mapped.brain")]
ggplot(subset(as.data.frame(colData(sce)), !is.na(mapped.brain)),
       aes(x = x_global_affine, 
           y = -y_global_affine)) + 
  geom_point(data = DF, colour = "grey",
             size = 0.1) +
  geom_point(aes(colour = mapped.brain),
             size = 1.5, alpha = 0.5) + 
  theme_classic() +
  facet_grid(mapped.brain ~ embryo + z) + 
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  coord_fixed() +
  NULL

gList = lapply(
  split.data.frame(as.data.frame(colData(sce_sub)),
                   interaction(sce_sub$embryo, sce_sub$z)), 
  function(df) {
    g = ggplot(df,
               aes(x = x_global_affine, 
                   y = -y_global_affine)) + 
      geom_point(data = subset(
        DF,
        # as.data.frame(colData(sce)),
        embryo %in% unique(df$embryo) &
          z %in% unique(df$z)), 
        colour = "grey",
        size = 0.1) +
      geom_point(
        colour = "black",
        # aes(colour = mapped.brain.score > 0.8),
        # aes(colour = cluster_brain),
        size = 1, alpha = 0.5) + 
      theme_classic() +
      facet_grid(embryo + z ~ mapped.brain) +
      # facet_grid(cut(as.numeric(mapped.brain.score),5) ~ mapped.brain) +
      guides(colour = guide_legend(override.aes = list(alpha = 1))) +
      coord_fixed() +
      xlab("") + 
      ylab("") + 
      theme(legend.position = "bottom") + 
      guides(colour = guide_legend(title = "",
                                   override.aes = list(size = 5, alpha = 1))) + 
      NULL
    return(g)
  })

gList[[1]]

lapply(gList, print)

sapply(names(gList), function(n){
  ggsave(gList[[n]], file = paste0("../Figures/brain_clusters_spatial_",
                                   n,
                                   ".pdf"),
         height = 6, width = 18)
})
```

# Match to brain subclusters in seqFISH

```{r}
table(sce_sub$cluster, sce_sub$cluster_brain)

table(sce_sub$cluster_sub, sce_sub$cluster_brain)

table(sce_sub$Neural, sce_sub$cluster_brain)

table(sce_sub$celltype_mapped_denoised, sce_sub$cluster_brain)
```

# Rename brain subclusters

Naming by Tim:

1. Tegmentum
2. Rhombencephalon 2
3. Rhombencephalon 3
4. Mesencephalon
5. Prosencephalon 2
6. Prosencephalon 1
7. Exclude
8. Rhombencephalon 1

the groups should be ordered in a certain way


8. Rhombencephalon 1
2. Rhombencephalon 2
3. Rhombencephalon 3
1. Tegmentum
4. Mesencephalon
6. Prosencephalon 1
5. Prosencephalon 2


```{r}
brain_sub_names = c(
  "1" = "Tegmentum",
  "2" = "Rhombencephalon 2",
  "3" = "Rhombencephalon 3",
  "4" = "Mesencephalon",
  "5" = "Prosencephalon 2",
  "6" = "Prosencephalon 1",
  "7" = NA,
  "8" = "Rhombencephalon 1"
)

brain_sub_names_order = c(
  "Rhombencephalon 1",
  "Rhombencephalon 2",
  "Rhombencephalon 3",
  "Tegmentum",
  "Mesencephalon",
  "Prosencephalon 1",
  "Prosencephalon 2"
)

sce$mapped.brain_named = brain_sub_names[as.character(sce$mapped.brain)]
table(sce$mapped.brain_named, useNA = "always")
names(sce$mapped.brain_named) <- colnames(sce)
```

# Output the cluster brain info

```{r}
saveRDS(sce$mapped.brain_named, file = "../analysis_output/E8.5/E8.5_mapped.brain_named.Rds")
```

# Cell-cell contact map of brain clusters

```{r}
graph = readRDS("../analysis_output/E8.5/E8.5_neighbourGraph_1.3.Rds")

nperm = 500

sce_sub <- sce[, !is.na(sce$mapped.brain_named)]

out_all = cellCellContact(sce_sub, group = "mapped.brain_named", graph, nperm = nperm)
g = cellCellContactMap(out_all)
g
cellCellContactMap(out_all, order = brain_sub_names_order)

split_categories = interaction(sce_sub$embryo, sce_sub$z)
sce_sub_split = sapply(unique(split_categories), function(s) {
  sce_sub[, split_categories == s]
})
names(sce_sub_split) <- as.character(unique(split_categories))

outList = lapply(sce_sub_split, cellCellContact, group = "mapped.brain_named", graph = graph, nperm = nperm, plot = FALSE)
gList = lapply(outList, cellCellContactMap, order = levels(cellCellContactMap(out_all)$data$subcluster_1))
# wrap_plots(gList)
gList2 = sapply(names(gList), function(n) gList[[n]] + ggtitle(n), simplify = FALSE)
g = wrap_plots(gList2)
# ggsave(g, file = "../Figures/brain_clusters_CCN_all.pdf", height = 18, width = 22)

out_mean = list("obs" = Reduce("+", lapply(outList, "[[", "obs"))/length(outList),
                "pmat" = Reduce("+", lapply(outList, "[[", "pmat"))/length(outList))
g = cellCellContactMap(out_mean, order = rev(brain_sub_names_order))
g

gg = as_ggplot(get_legend(g))
ggsave(gg, file = "../Figures/cellcellcontact_legend.pdf", 
       height = 1.5, width = 6)

brain_cols = readRDS("../analysis_output/E8.5/E8.5_brain_subcluster_colours.Rds")

if (FALSE) {
  h = cellCellContactHeatmap(out_mean, col = brain_cols, split_n = 4)
  h
  
  h = cellCellContactHeatmapTriangle(out_mean, col_ann = brain_cols, split_n = 4)
  h
  pdf(file = "../Figures/mapped.brain_named_cellcellcontact.pdf",
      height = 8, width = 8)
  print(h)
  dev.off()
}

h = cellCellContactHeatmap(out_mean, col = brain_cols, order = brain_sub_names_order)
h

h = cellCellContactHeatmapTriangle(out_mean, col_ann = brain_cols, order = rev(brain_sub_names_order))
h

pdf(file = "../Figures/mapped.brain_named_cellcellcontact.pdf",
    height = 8, width = 8)
print(h)
dev.off()

```

# Finish

```{r}
sessionInfo()
```