---
title: "E8.5 seqFISH - Downstream mesoderm analysis using Han et al Data"
author: "Shila Ghazanfar"
date: "17/08/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 16
                     fig_height: 12
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, cache.lazy = FALSE)
```

```{r}
set.seed(2020)
```

```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
library(ggmap)
library(randomForest)
library(batchelor)
library(igraph)
library(cowplot)
library(ggpubr)
library(Matrix)
library(grDevices)
library(RColorBrewer)
```

```{r}
sapply(list.files("../functions/", full.names = TRUE), source)
```

```{r}
na_colour = "white"
polygon_size = 0.1
polygon_colour = "black"
```

# Summary

Here we examine the mesodermal layers that are around the gut tube, and see
how they relate in terms of their gene expression patterns in space.

# Load spatial data

```{r}
sce = readRDS("../analysis_output/E8.5/E8.5_sce_filt_unlabelled.Rds")
sce
sce <- reNormalise(sce)
```

# Add cell type and clustering information

```{r}
mapping_dt = readRDS("../analysis_output/E8.5/celltype_annotation_refined.Rds")
colData(sce) <- cbind(colData(sce), mapping_dt[colnames(sce), setdiff(colnames(mapping_dt),colnames(colData(sce)))])
```

# Also add Nowotschin annotation

```{r}
now_joint_df = readRDS("../analysis_output/E8.5/E8.5_Nowotschin_joint_mnn_df.Rds")
sce$gut_tube_subtype = now_joint_df[colnames(sce), "ctype"]
colData(sce)[sce$celltype_mapped_refined != "Gut tube","gut_tube_subtype"] <- NA
```

# And Han annotation

Only for those cell types listed

```{r}
meso_types = c("Anterior somitic tissues",
               "Cardiomyocytes",
               "Cranial mesoderm",
               "Endothelium",
               "Gut tube",
               "Mesenchyme",
               "Splanchnic mesoderm",
               "Sclerotome")

han_joint_df = readRDS("../analysis_output/E8.5/E8.5_Han_joint_mnn_df.Rds")
sce$han_subtype = han_joint_df[colnames(sce), "ctype"]
colData(sce)[!sce$celltype_mapped_refined %in% meso_types,"han_subtype"] <- NA
```

# Add cell neighbourhood

```{r}
graph = readRDS("../analysis_output/E8.5/E8.5_neighbourGraph_1.3.Rds")
```

# expand network to the gut tube cells

```{r}
gut_tube_cells = colnames(sce)[sce$celltype_mapped_refined == "Gut tube"]

expanded_vertices_list = neighborhood(graph, order = 3, nodes = which(V(graph)$name %in% gut_tube_cells))
expanded_vertices = sort(unique(unlist(lapply(expanded_vertices_list, function(x) x$name))))
length(expanded_vertices)
length(setdiff(expanded_vertices, gut_tube_cells))
nearby_meso_cells = intersect(expanded_vertices, colnames(sce)[sce$celltype_mapped_refined %in% setdiff(meso_types, "Gut tube")])
length(nearby_meso_cells)

sort(table(colData(sce)[nearby_meso_cells, "celltype_mapped_refined"], useNA = "always"))
sort(table(colData(sce)[nearby_meso_cells, "gut_tube_subtype"], useNA = "always"))
sort(table(colData(sce)[nearby_meso_cells, "han_subtype"], useNA = "always"))
```

# plot these nearby cells' Han labels

```{r}
colourCount = length(unique(han_joint_df$ctype))
getPalette = colorRampPalette(RColorBrewer::brewer.pal(12, "Accent"))
han_cols = getPalette(colourCount)
names(han_cols) = levels(han_joint_df$ctype)
```

Figure panel for MS:

```{r}
sce$han_subtype_mask = sce$han_subtype
sce$han_subtype_mask[!colnames(sce) %in% nearby_meso_cells] <- NA
sce$han_subtype_mask[substring(sce$han_subtype_mask,1,1) == "E"] <- NA

g = ggplot(as.data.frame(colData(sce)), aes(x = x_global_affine, y = -y_global_affine)) + 
  geom_point(aes(colour = han_subtype_mask), size = 0.5) + 
  facet_wrap(z ~ embryo) +
  theme_classic() +
  scale_colour_manual(values = han_cols, na.value = "lightgrey") + 
  guides(colour = guide_legend(title = "", override.aes = list(size = 5))) +
  coord_fixed() +
  NULL
g

# make graph just for embryo1 z 2 

bdf = getSegmentationVerticesDF(colData(sce),
                                othercols = c("uniqueID", "z", "embryo", "han_subtype_mask"))

g = ggplot(
  subset(bdf, embryo == "embryo1" & z == 2), 
  aes(x = segmentation_vertices_x_global, y = -segmentation_vertices_y_global)) +
  geom_polygon(aes(group = uniqueID, fill = han_subtype_mask),
               colour = "black", size = 0.1) + 
  theme_classic() +
  scale_fill_manual(values = han_cols, na.value = "lightgrey") + 
  guides(fill = guide_legend(title = "", override.aes = list(size = 5))) +
  coord_fixed() +
  xlab("") + 
  ylab("") +
  theme(axis.line = element_blank()) + 
  theme(axis.text = element_blank()) + 
  theme(axis.ticks = element_blank()) + 
  NULL

ggsave(g + theme(legend.position = "none"),
       file = "../Figures/Han_mesodermal_embryo1.pdf", height = 10, width = 8)

gg = as_ggplot(get_legend(g + theme(legend.position = "right") + 
                            guides(fill = guide_legend(title = ""))))
print(gg)

ggsave(gg,  file = "../Figures/Han_mesodermal_legend.pdf",
       height = 4, width = 3)

```

# cell separability of these subtypes

```{r}
library(e1071)
sce_sub = sce[, !is.na(sce$han_subtype_mask)]
dist_sep = cellCellSeparability(sce_sub, split = interaction(sce_sub$embryo, sce_sub$z),
                                group = "han_subtype_mask",
                                coordNames = c("x_global_affine", 
                                               "y_global_affine"),
                                plot = FALSE)

g = cellCellSeparabilityMap(dist_sep, order = mixedsort(unique(sce_sub$han_subtype_mask)))
g
```

# Calculate principal curves using virtual dissection

```{r}
gut_dissection = readRDS("../analysis_output/E8.5/gut_virtual_dissection.Rds")
gut_dissection

DV_class_list = list()

# first identify cells that are in the foregut region of each embryo
# then identify cells that are in the dorsal foregut region of each embryo
for (embryoval in sort(unique(sce$embryo))) {
  
  sce_sub = sce[, sce$embryo == embryoval]
  
  in_foregut = pracma::inpolygon(sce_sub$x_global_affine, -sce_sub$y_global_affine,
                                 gut_dissection[["foregut"]][[embryoval]][["x"]],
                                 gut_dissection[["foregut"]][[embryoval]][["y"]])
  
  in_dorsal_foregut = pracma::inpolygon(sce_sub$x_global_affine, -sce_sub$y_global_affine,
                                        gut_dissection[["dorsal_foregut"]][[embryoval]][["x"]],
                                        gut_dissection[["dorsal_foregut"]][[embryoval]][["y"]])
  
  par(mfrow = c(1,2))
  plot(sce_sub$x_global_affine, -sce_sub$y_global_affine, pch = 16, cex = 0.5, asp = 1,
       col = ifelse(sce_sub$celltype_mapped_refined == "Gut tube" & in_dorsal_foregut & in_foregut, "red", "black"),
       main = embryoval)
  
  plot(sce_sub$x_global_affine, -sce_sub$y_global_affine, pch = 16, cex = 0.5, asp = 1,
       col = ifelse(sce_sub$celltype_mapped_refined == "Gut tube" & !in_dorsal_foregut & in_foregut, "red", "black"),
       main = embryoval)
  
  names(in_dorsal_foregut) <- colnames(sce_sub)
  names(in_foregut) <- colnames(sce_sub)
  
  DV_class_sub = rep(NA, ncol(sce_sub))
  names(DV_class_sub) <- colnames(sce_sub)
  DV_class_sub[sce_sub$celltype_mapped_refined == "Gut tube" & in_dorsal_foregut & in_foregut] <- "Dorsal"
  DV_class_sub[sce_sub$celltype_mapped_refined == "Gut tube" & !in_dorsal_foregut & in_foregut] <- "Ventral"
  table(DV_class_sub)
  
  par(mfrow = c(1,1))
  plot(sce_sub$x_global_affine, -sce_sub$y_global_affine, pch = 16, cex = 0.5, asp = 1,
       col = c("Dorsal" = "red", "Ventral" = "blue")[DV_class_sub],
       main = embryoval)
  
  DV_class_list[[embryoval]] <- DV_class_sub
  
}

names(DV_class_list) <- NULL
DV_class = unlist(DV_class_list)[colnames(sce)]

sce$DV_class <- DV_class
```

# identify AP using principal curves

```{r}
library(princurve)

sce_split = splitSCE(sce, interaction(sce$DV_class, sce$embryo))

fitList = list()
lengthList = list()

for (i in 1:length(sce_split)) {
  
  print(i)
  
  if (ncol(sce_split[[i]]) == 0) next
  
  print(names(sce_split)[i])
  
  coords = as.matrix(colData(sce_split[[i]])[, c("x_global_affine", "y_global_affine")])
  coords[,2] <- -coords[,2]
  
  plot(coords, asp = 1, pch = 16)
  
  fit0 = principal_curve(coords, df = 6)
  
  points(fit0, col = "blue")
  
  text(coords, label = fit0$ord)
  lines(coords[fit0$ord,], type = "l")
  ordered_cells = rownames(coords[fit0$ord,])
  
  # make sure order goes from anterior to posterior
  if (diff(sce[,c(ordered_cells[c(1,length(ordered_cells))])]$y_global_affine) < 0) {
    reordered_cells <- rev(ordered_cells)
    cells_length = -fit0$lambda[reordered_cells] + max(fit0$lambda[reordered_cells])
  } else {
    reordered_cells = ordered_cells
    cells_length = fit0$lambda[reordered_cells]
  }
  
  # this should be zero
  print(cells_length[reordered_cells[1]])
  
  plot(coords, asp = 1, pch = 16, cex = 0.5, main = i)
  points(fit0, col = "blue", cex = 0.1)
  points(coords[reordered_cells[1], , drop = FALSE], col = "orange", cex = 5, pch = 16)
  lines(coords[reordered_cells,], type = "l", col = "orange")
  
  fitList[[names(sce_split)[i]]] <- fit0
  
  plot(sce[,reordered_cells]$y_global_affine, main = i)
  plot(cells_length, main = i)
  
  lengthList[[names(sce_split)[i]]] <- cells_length
  
}

names(lengthList) <- NULL

AP_value = unlist(lengthList)[colnames(sce)]
names(AP_value) <- colnames(sce)

sce$AP_value <- AP_value
```

# save output

```{r}
DV_vals = list(DV_class = sce$DV_class, AP_value = sce$AP_value)
names(DV_vals[["DV_class"]]) <- colnames(sce)
names(DV_vals[["AP_value"]]) <- colnames(sce)
saveRDS(DV_vals,
        file = "../analysis_output/E8.5/gut_tube_DV_split.Rds")
saveRDS(fitList, file = "../analysis_output/E8.5/gut_tube_DV_princurve.Rds")
```

# Finish

```{r}
sessionInfo()
```