---
title: "E8.5 seqFISH - Mesenchyme"
author: "Shila Ghazanfar"
date: "27/10/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 16
                     fig_height: 12
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, cache.lazy = FALSE)
```

```{r}
set.seed(2020)
```

```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
library(ggmap)
library(randomForest)
library(batchelor)
library(igraph)
library(cowplot)
library(ggpubr)
library(Matrix)
library(grDevices)
library(RColorBrewer)
```

```{r}
library(SpatialUtils)
```

```{r}
na_colour = "white"
polygon_size = 0.1
polygon_colour = "black"
```

# Summary

Here we examine the mesenchyme.

# Load spatial data

```{r}
sce = readRDS("../analysis_output/E8.5/E8.5_sce_filt_unlabelled.Rds")
sce

sce <- reNormalise(sce)
```

# Add cell type and clustering information

```{r}
mapping_dt = readRDS("../analysis_output/E8.5/celltype_annotation_refined.Rds")
colData(sce) <- cbind(colData(sce), mapping_dt[colnames(sce), setdiff(colnames(mapping_dt),colnames(colData(sce)))])
```

# Add cell neighbourhood

```{r}
graph = readRDS("../analysis_output/E8.5/E8.5_neighbourGraph_1.3.Rds")
```

# Load imputed data

```{r}
library(DelayedArray)
library(HDF5Array)

file_combined = "../analysis_output/imputation/combined.h5"
cnames = readRDS("../analysis_output/imputation/combined_cnames.Rds")
rnames = readRDS("../analysis_output/imputation/combined_rnames.Rds")

imp <- HDF5Array(filepath = file_combined, name = "logcounts")
rownames(imp) <- rnames
colnames(imp) <- cnames

genes_imp = rnames
```

# Load mouse gastrulation metadata

```{r}
atlas = readRDS("../analysis_output/E8.5/atlas_for_celltype_mapping.Rds")
atlas_umap = reducedDim(atlas, "umap")

plot(atlas_umap, pch = 16, cex = 0.3, asp = 1)
```

Plot the relative weighting of the mesenchyme cells for each of the mesenchyme 
subregions. For each M-group, calculate the sum of ranking for these cells

# Do the mesenchyme cells cluster separately?

```{r}
# perform louvain clustering on the corrected mes
if (!file.exists("../analysis_output/E8.5/Mesenchyme_subclusters.Rds")) {
  snn = buildSNNGraph(t(mes_corrected))
  clusters_graph = cluster_louvain(snn)
  clusters = as.numeric(membership(clusters_graph))
  names(clusters) <- rownames(mes_corrected)
  table(clusters, useNA = "always")
  saveRDS(clusters, file = "../analysis_output/E8.5/Mesenchyme_subclusters.Rds")
}
clusters = readRDS("../analysis_output/E8.5/Mesenchyme_subclusters.Rds")

mes_umap_df_cluster = mes_umap_df
mes_umap_df_cluster$cluster = factor(clusters[as.character(mes_umap_df_cluster$uniqueID)])

g_umap = ggplot(mes_umap_df_cluster,
                aes(x = UMAP1, y = UMAP2)) + 
  geom_point(aes(colour = cluster), size = 1) +
  theme_classic() +
  coord_fixed() +
  theme(axis.ticks = element_blank()) + 
  theme(axis.text = element_blank()) +
  scale_colour_manual(values = M_colours_2) +
  guides(colour = guide_legend(title = "Cluster", override.aes = list(size = 5))) +
  NULL
g_umap + facet_grid(~embryo)
ggsave(g_umap, file = "../Figures/Mesenchyme_subclusters_umap.pdf",
       height = 6, width = 7)
ggsave(g_umap + facet_grid(~embryo), file = "../Figures/Mesenchyme_subclusters_umap_splitembryo.pdf",
       height = 6, width = 7)
g_leg = as_ggplot(get_legend(g_umap))
g_leg
ggsave(g_leg, file = "../Figures/Mesenchyme_subclusters_legend.pdf",
       height = 3, width = 2)

g_space = ggplot(mes_umap_df_cluster,
                aes(x = x, y = -y)) + 
  geom_point(data = data.frame(x = sce$x_global_affine, y = sce$y_global_affine, embryo = sce$embryo), colour = "lightgrey", size = 0.5) +
  geom_point(aes(colour = cluster), size = 1) +
  theme_classic() +
  facet_grid(~embryo) + 
  guides(colour = guide_legend(title = "", override.aes = list(size = 5))) +
  xlab("") +
  ylab("") +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  coord_fixed() +
  NULL
g_space

g_space + g_umap + plot_layout(nrow = 2, heights = c(2.5,2))

addmargins(table(mes_umap_df_cluster$mes, mes_umap_df_cluster$cluster))
```

Plot the relative weighting of the mesenchyme cells for each of the mesenchyme 
subclusters. For each cluster, calculate the sum of ranking for these cells

```{r}
k.mapped = readRDS("../analysis_output/E8.5/E8.5_naive_k.mapped_libsize.Rds")

gList = list()

subregions = sort(unique(as.character(clusters)))

for (subregion in subregions) {
  k.mapped_sub = k.mapped[names(which(clusters == subregion)),]
  k.mapped_long = c(k.mapped_sub)
  k.mapped_longvalues = rep(ncol(k.mapped_sub):1, each = nrow(k.mapped_sub))
  # k.mapped_longvalues = rep(c(1,rep(0,ncol(k.mapped_sub)-1)), each = nrow(k.mapped_sub))
  subregion_score = tapply(k.mapped_longvalues, k.mapped_long, sum)/nrow(k.mapped_sub)
  
  subregion_score_all = rep(0, nrow(atlas_umap))
  names(subregion_score_all) <- rownames(atlas_umap)
  subregion_score_all[names(subregion_score)] <- subregion_score
  
  hist(subregion_score_all)
  
  df = data.frame(
    cell = rownames(atlas_umap),
    UMAP1 = atlas_umap[,1],
    UMAP2 = atlas_umap[,2],
    subregion_score = subregion_score_all
  )
  
  gList[[subregion]] <- ggplot(sort_df(df, "subregion_score"),
                               aes(x = UMAP1, y = UMAP2)) + 
    geom_point(aes(colour = subregion_score), size = 0.3) +
    coord_fixed() +
    theme_classic() +
    scale_colour_gradient(low = "grey", high = "black") +
    theme(axis.text = element_blank()) +
    theme(axis.ticks = element_blank()) +
    guides(colour = guide_colourbar(title = "")) +
    ggtitle(subregion) +
    NULL
}

g = wrap_plots(gList, nrow = 1)
g

ggsave(g, file = "../Figures/Mesenchyme_subclusters_10X_umap.pdf",
       height = 5, width = 15)

```


Find marker genes for these clusters

```{r}
sce_sub$mes_cluster = paste0("cluster_", clusters[colnames(sce_sub)])

marker_array = getMarkerArray(sce_sub, 
                              group = "mes_cluster",
                              block = "embryo",
                              pseudobulk = FALSE)


discriminating_array_list = list()

groupValue = "mes_cluster"

groups = setdiff(sort(unique(colData(sce_sub)[, groupValue])), "")

for (grouplevel in groups) {
    
    print(grouplevel)
    
    groups2 = setdiff(groups, grouplevel)
    
    for (grouplevel2 in groups2) {
        
        print(grouplevel2)
        
        ma = getMarkerArray(sce_sub,
                            # group = "iterMNN_stable",
                            group = groupValue,
                            block = "embryo",
                            subset = colData(sce_sub)[, groupValue] %in% c(grouplevel,
                                                               grouplevel2),
                            pseudobulk = FALSE,
                            verbose = FALSE)
        
        discriminating_array_list[[grouplevel]][[grouplevel2]] <- ma
        
    }
    
}

discriminating_array_list[["cluster_5"]][["cluster_3"]]["Postn",,]
marker_array["Postn",,]


markerArrayPlot(marker_array,
                grouplevel = "cluster_5",
                otherGroupLabel = "All other cells",
                onlyLabelTopRanked = TRUE,
                FDR = 0.05,
                LFC = 0.5)

sig_genes = sort(unique(rownames(which(marker_array[,,"FDR"] < 0.05 & abs(marker_array[,,"LFC"]) > 0.5, arr = TRUE))))
length(sig_genes)

meanExp = sumCountsAcrossCells(logcounts(sce_sub)[sig_genes,],
                               droplevels(interaction(sce_sub$embryo, sce_sub$mes_cluster)),
                               average = TRUE)
meanExp_rel = t(apply(meanExp, 1, function(x)x/max(x)))
dim(meanExp)
dim(meanExp_rel)
h = Heatmap(meanExp_rel, col = c("grey75", "cornflowerblue", "black"),
        column_split = gsub("embryo[0-9]\\.", "", colnames(meanExp_rel))
        ,
        row_split = 5
        )
h

pdf(file = "../Figures/Mesenchyme_subclusters_markers.pdf",
    height = 15, width = 10)
print(h)
dev.off()
```

Find imputed marker genes for these cells

```{r}
sce_sub_imp = SingleCellExperiment(
  assays = list(logcounts = imp[,colnames(sce_sub)])
)
sce_sub_imp$mes_cluster = sce_sub$mes_cluster
sce_sub_imp$embryo = sce_sub$embryo

marker_array = getMarkerArray(sce_sub_imp, 
                              # group = "iterMNN_stable",
                              group = "mes_cluster",
                              # subset = sce_sub$gut_tube_subtype != "Large intestine/Colon",
                              block = "embryo",
                              pseudobulk = FALSE)


discriminating_array_list = list()

groupValue = "mes_cluster"

groups = setdiff(sort(unique(colData(sce_sub_imp)[, groupValue])), "")

for (grouplevel in groups) {
    
    print(grouplevel)
    
    groups2 = setdiff(groups, grouplevel)
    
    for (grouplevel2 in groups2) {
        
        print(grouplevel2)
        
        ma = getMarkerArray(sce_sub_imp,
                            group = groupValue,
                            block = "embryo",
                            subset = colData(sce_sub_imp)[, groupValue] %in% c(grouplevel,
                                                               grouplevel2),
                            pseudobulk = FALSE,
                            verbose = FALSE)
        
        discriminating_array_list[[grouplevel]][[grouplevel2]] <- ma
        
    }
    
}

discriminating_array_list[["cluster_5"]][["cluster_3"]]["Postn",,]
marker_array["Postn",,]

markerArrayPlot(marker_array,
                grouplevel = "cluster_5",
                otherGroupLabel = "All other cells",
                onlyLabelTopRanked = TRUE,
                FDR = 0.05,
                LFC = 0.5)

sig_genes = sort(unique(rownames(which(marker_array[,,"FDR"] < 0.05 & abs(marker_array[,,"LFC"]) > 0.5, arr = TRUE))))
length(sig_genes)
table(sig_genes %in% rownames(sce))
sort(setdiff(sig_genes, rownames(sce)))

meanExp = sumCountsAcrossCells(logcounts(sce_sub_imp)[sig_genes,],
                               droplevels(interaction(sce_sub_imp$embryo, sce_sub_imp$mes_cluster)),
                               average = TRUE)
meanExp_rel = t(apply(meanExp, 1, function(x)x/max(x)))
dim(meanExp)
dim(meanExp_rel)
h = Heatmap(meanExp_rel, col = c("grey75", "cornflowerblue", "black"),
            column_split = gsub("embryo[0-9]\\.", "", colnames(meanExp_rel)),
            row_split = 5)

pdf(file = "../Figures/Mesenchyme_subclusters_markers_imputed.pdf",
    height = 25, width = 10)
print(h)
dev.off()
```

# make subclusters by subsetting the batch corrected pca

```{r}
snn = buildSNNGraph(sce_sub, use.dimred = "mbPCA_cosine_MNN")
clusters_graph = cluster_louvain(snn)
clusters_all = paste0("cluster_all", as.numeric(membership(clusters_graph)))
names(clusters_all) <- sce_sub$uniqueID

table(clusters_all)

table(clusters_all, clusters)

mes_umap_df_cluster$cluster_all = clusters_all[as.character(mes_umap_df_cluster$uniqueID)]

g_space = ggplot(mes_umap_df_cluster,
                aes(x = x, y = -y)) + 
  geom_point(data = data.frame(x = sce$x_global_affine, y = sce$y_global_affine, embryo = sce$embryo), colour = "lightgrey", size = 0.5) +
  geom_point(aes(colour = cluster_all), size = 1) +
  theme_classic() +
  facet_grid(~embryo) + 
  guides(colour = guide_legend(title = "", override.aes = list(size = 5))) +
  xlab("") +
  ylab("") +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  coord_fixed() +
  NULL
g_space
```

spatial segmented plot of mesenchyme clusters

```{r}
sce$mes_cluster = rep(NA, ncol(sce))
sce[,colnames(sce_sub)]$mes_cluster <- sce_sub$mes_cluster

boundary_polygons$mes_cluster = sce[,as.character(boundary_polygons$uniqueID)]$mes_cluster

M_colours_3 = M_colours_2
names(M_colours_3) <- paste0("cluster_", 1:5)

colval = "lightgrey"

for (embryoval in c("embryo1","embryo2","embryo3")) {
  
  g = ggplot() +
    geom_polygon(aes(x = segmentation_vertices_x_global_affine,
                     y = -segmentation_vertices_y_global_affine,
                     group = uniqueID,
                     fill = mes_cluster,
                     colour = !is.na(mes_cluster)
                     ),
                 size = 0.1,
                 data = subset(boundary_polygons, embryo == embryoval & z == 2)) +
    scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "lightgrey")) +
    facet_grid(z~embryo) +
    scale_fill_manual(values = M_colours_3) +
    theme_classic() +
    coord_fixed() +
    theme(legend.position = "none") + 
    ggtitle("") +
    xlab("") +
    theme(axis.line = element_blank()) + 
    theme(axis.text = element_blank()) + 
    theme(axis.title.y = element_blank()) +
    theme(axis.ticks = element_blank()) +
    theme(strip.background = element_blank()) +
    theme(strip.text = element_blank()) +
    theme(plot.title = element_text(hjust = 0.5, size = 15)) +
    add_scalebar() +
    NULL
  # g
  ggsave(g, file = paste0("../Figures/Mesenchyme_subclusters_spatial_", embryoval, ".pdf"),
         height = 10, width = 8)
}

gg = as_ggplot(get_legend(g +
                            theme(legend.position = "right") + 
                            guides(fill = guide_legend(title = ""))))
gg
ggsave(gg, file = "../Figures/Mesenchyme_subclusters_spatial_legend.pdf",
       height = 4, width = 2)
```

# proportional barplot of the neighbouring cell types of the different anatomical subregions

for each mesenchyme cell group, identify the set of neighbours, then plot
their breakdown in terms of celltype

```{r}
subregions = sort(unique(as.character(sce_sub$mes_cluster)))

neighbour_df = NULL

for (subregion in subregions) {
  
  subregion_cells = colnames(sce[,sce$mes_cluster == subregion & !is.na(sce$mes_cluster)])
  
  expanded_vertices_list = neighborhood(graph, order = 1, nodes = which(V(graph)$name %in% subregion_cells))
  expanded_vertices = sort(setdiff(unique(unlist(lapply(expanded_vertices_list, function(x) x$name))), subregion_cells))
  length(expanded_vertices)

  df_sub = data.frame(uniqueID = expanded_vertices,
                                   celltype = sce[,expanded_vertices]$celltype_mapped_refined,
                                   subregion = subregion)
  mask_types = c(names(which(table(df_sub$celltype)/nrow(df_sub) < 0.05)), "Low quality")
  df_sub$celltype_mask = as.character(df_sub$celltype)
  df_sub$celltype_mask[df_sub$celltype_mask %in% mask_types] <- "Other"
  
  neighbour_df <- rbind(neighbour_df,
                        df_sub)
  
}

neighbour_df$embryo = sce[,as.character(neighbour_df$uniqueID)]$embryo
dim(neighbour_df)
table(neighbour_df$subregion)

g_all = ggplot(subset(neighbour_df, celltype != "Low quality"), aes(x = subregion)) + 
  geom_bar(aes(fill = celltype, colour = celltype), position = "fill") + 
  theme_classic() +
  scale_fill_manual(values = celltype_colours, aesthetics = c("fill", "colour")) +
  theme(legend.position = "none") +
  xlab("") + 
  ylab("") + 
  NULL
g_all

g = ggplot(neighbour_df, aes(x = embryo)) + 
  geom_bar(aes(fill = celltype_mask, colour = celltype_mask), colour = "black", position = "fill") + 
  theme_classic() +
  scale_fill_manual(values = c(celltype_colours, "Other" = "grey"), aesthetics = c("fill", "colour")) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  xlab("") + 
  ylab("") + 
  facet_grid(~subregion) +
  scale_y_continuous(limits = c(0,1), expand = c(0.01,0.01)) +
  NULL
g
ggsave(g, file = "../Figures/Mesenchyme_subclusters_neighbour_proportions.pdf",
       height = 6, width = 4)

g_all + g
```

# Violin plots of expression for given genes

```{r}
genes = c("Tbx18", "Wt1", "Msx1", "Msx2", "Id1", "Gata5", "Sfrp1", "Gata4", "Gata6", "Sfrp5",
          "Flrt3","Shisa2")

for (gene in genes) {
mes_umap_df_gene = mes_umap_df_cluster

if (gene %in% rownames(sce)) {
  suffix = ""
  mes_umap_df_gene$gene <- logcounts(sce)[gene,as.character(mes_umap_df_gene$uniqueID)]
} else {
  suffix = " (imputed)"
  mes_umap_df_gene$gene <- imp[gene,as.character(mes_umap_df_gene$uniqueID)]  
}

g_vio = ggplot(mes_umap_df_gene,
       aes(x = embryo, y = gene)) + 
  geom_violin(aes(fill = cluster), scale = "width", draw_quantiles = 0.5) + 
  theme_classic() +
  facet_grid(~cluster) + 
  scale_fill_manual(values = M_colours_2) +
  ylab("Expression") +
  guides(fill = guide_legend(title = "")) +
  xlab("") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
    theme(strip.background = element_rect(colour = "white")) +
  theme(strip.text = element_blank()) +
  theme(legend.position = "none") +
  NULL
g_vio

g_umap = ggplot(mes_umap_df_gene,
                aes(x = UMAP1, y = UMAP2)) + 
  geom_point(aes(colour = gene), size = 1) +
  theme_classic() +
  scale_colour_gradient2(low = "gray75", mid = "cornflowerblue", high = "black",
                         midpoint = 0.5*max(mes_umap_df_gene$gene)) +
  ylab("UMAP2") +
  guides(colour = guide_colourbar(title = "")) +
  xlab("UMAP1") +
   theme(axis.text = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(legend.position = "none") +
  coord_fixed() +
  NULL
g_umap
g_umap + facet_grid(~embryo)

g_space = ggplot(mes_umap_df_gene,
                aes(x = x, y = -y)) + 
  geom_point(data = data.frame(x = sce$x_global_affine, y = sce$y_global_affine, embryo = sce$embryo), colour = "lightgrey", size = 0.5) +
  geom_point(aes(colour = gene)) +
  theme_classic() +
  facet_grid(~embryo) + 
  scale_colour_gradient2(low = "gray75", mid = "cornflowerblue", high = "black",
                         midpoint = 0.5*max(mes_umap_df_gene$gene)) +
  guides(colour = guide_colourbar(title = "")) +
  xlab("") +
  ylab("") +
  theme(legend.position = "none") +
  coord_fixed() +
  theme(axis.text = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(axis.line = element_blank()) +
  theme(strip.background = element_rect(colour = "white")) +
  theme(strip.text = element_blank()) +
  add_scalebar() +
  NULL
g_space

g = g_space + g_umap + g_vio + plot_layout(nrow = 1, widths = c(4,1,1.5))

ggsave(g, file = paste0("../Figures/Mesenchyme_subclusters_expression_", gene,".pdf"),
       height = 5, width = 15)
}
```

# GO analysis of the mesenchyme cluster marker genes

```{r}
marker_genes_list = sapply(dimnames(marker_array)[[2]], function(x){
  df = as.data.frame(marker_array[,x,])
  genes = rownames(subset(df, FDR < 0.2 & LFC >= 0.2))
  return(genes)
}, simplify = FALSE)
lapply(marker_genes_list, length)


genesetGOtest = function(set, universe, termsList) {
  # set is the character vector of genes in geneset
  # universe is character vector of genes to include in universe
  # termsList is a list of the pathways
  
  termsListfiltered = lapply(termsList, function(x) return(intersect(universe, x)))
  keepForTesting = unlist(lapply(termsListfiltered, length)) >= 8 & unlist(lapply(termsListfiltered, length)) <= 500
  
  pval = sapply(1:length(termsList), function(i){
    
    if (!keepForTesting[i]) return(1)
    
    termGenes = termsListfiltered[[i]]
    return(fisher.test(table(universe %in% set, universe %in% termGenes), alt = "g")$p.value)
  })
  names(pval) <- names(termsListfiltered)
  return(pval)
}

GO_list = readRDS("../analysis_output/E8.5/GO_list.Rds")

GO_list_sig = GO_list[unlist(lapply(GO_list, function(x) any(x %in% unlist(marker_genes_list))))]
length(GO_list_sig)

GO_testing_genes = marker_genes_list
length(GO_testing_genes)
head(names(GO_testing_genes))
GO_res_list = lapply(GO_testing_genes, function(set){
  genesetGOtest(set, rnames, GO_list_sig)
})


ntop = 20
library(stringr)
gList_GO = lapply(GO_res_list, function(pval) {
  df = data.frame(term = factor(names(pval), levels = c(names(pval), "")),
                  pval = pval,
                  qval = p.adjust(pval, method = "BH"))
  df$label = df$term
  df$label[pval != 1] <- ""
  df_sorted = sort_df(df, "pval")[1:ntop,]
  df_sorted$term = factor(df_sorted$term, levels =  rev(df_sorted$term))
  
  g = ggplot(df_sorted, aes(x = term, y = -log10(pval), fill = qval < 0.05)) +
    theme_classic() +
    geom_col() +
    coord_flip() +
    xlab("") +
    ylab(expression("-log10(P-value)")) +
    # geom_hline(yintercept = -log10(0.01), colour = "red", linetype = "dashed", size = 1.2) +
    scale_fill_manual(values = c("TRUE" = "dimgrey", "FALSE" = "peachpuff")) +
    theme(legend.position = "none") +
    theme(axis.text.y = element_text(size = 10)) +
    theme(title = element_text(hjust = 0.5)) +
    scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) +
    NULL
  return(g)
  
})
g = wrap_plots(sapply(names(gList_GO), function(x) gList_GO[[x]] + ggtitle(x), simplify = FALSE), nrow = 1)
ggsave(g, file = "../Figures/Mesenchyme_subclusters_GO.pdf", height = 8, width = 30)
gList_GO[[2]] + gList_GO[[3]]

# dot plot of all terms
GO_df = do.call(rbind, sapply(1:length(GO_res_list), function(i) {
  pval = GO_res_list[[i]]
  sample = paste0("cluster ",i)
  if (i == 11) {
    sample = "all"
  }
  df = data.frame(term = factor(names(pval), levels = c(names(pval), "")),
                  pval = pval,
                  qval = p.adjust(pval, method = "BH"))
  df$label = df$term
  df$label[pval != 1] <- ""
  df_sorted = sort_df(df, "pval")#[1:ntop,]
  df_sorted$rank = rank(df_sorted$pval)
  df_sorted$term = factor(df_sorted$term, levels =  rev(df_sorted$term))
  df_sorted$cluster = sample
  return(df_sorted)
}, simplify = FALSE))

any_sig_terms = sort(unique(subset(GO_df, rank <= 20 & qval < 0.05 & cluster != "all")$term))
length(any_sig_terms)

GO_df_sub = subset(GO_df, term %in% any_sig_terms)

GO_sub_raw = cast(GO_df_sub, term ~ cluster, value = "pval")
GO_sub = -log10(as.matrix(GO_sub_raw[,-1]))
colnames(GO_sub) <- colnames(GO_sub_raw)[2:ncol(GO_sub_raw)]
rownames(GO_sub) <- GO_sub_raw[,1]

order_terms = rownames(GO_sub)[hclust(dist(GO_sub))$order]
order_clusters = colnames(GO_sub)[hclust(dist(t(GO_sub)))$order]

GO_df_sub$term <- factor(GO_df_sub$term,
                         levels = rev(order_terms))
GO_df_sub$cluster <- factor(GO_df_sub$cluster,
                         levels = order_clusters)

g = ggplot(GO_df_sub, aes(x = cluster, y = term)) + 
  geom_point(aes(size = -log10(pval), colour = qval < 0.05)) + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  scale_colour_manual(values = c("FALSE" = "grey", "TRUE" = "black"),
                      labels = c("FDR-adjusted\nP-value > 0.05",
                                 "FDR-adjusted\nP-value < 0.05")) +
  guides(colour = guide_legend(title = "")) +
  guides(size = guide_legend(title = "-log10(P-value)")) +
  theme(legend.position = "right") +
  xlab("") +
  ylab("") +
  NULL
g
ggsave(g, file = "../Figures/Mesenchyme_GO_heatmap.pdf",
       height = 15, width = 10)

bmp_genes = sort(unlist(GO_list_sig[names(GO_list_sig)[grepl("bmp", tolower(names(GO_list_sig)))]]))
lapply(GO_testing_genes, function(x) intersect(x, bmp_genes))

fgf_genes = sort(unlist(GO_list_sig[names(GO_list_sig)[grepl("fibroblast gro", tolower(names(GO_list_sig)))]]))
lapply(GO_testing_genes, function(x) intersect(x, fgf_genes))


cushion_genes = sort(unlist(GO_list_sig[names(GO_list_sig)[grepl("endocardial cushion formation", tolower(names(GO_list_sig)))]]))
lapply(GO_testing_genes, function(x) intersect(x, cushion_genes))
```

# Finish

```{r}
sessionInfo()
```