---
title: "E8.5 seqFISH - Integrating with atlas data"
author: "Shila Ghazanfar"
date: "23/07/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 10 
                     fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE, cache.lazy = FALSE)
```

```{r}
set.seed(2020)
```

# Summary

In this document I focus on classification of cell types using MNN using the 
gastrulation atlas data. Using library size normalised and z-stack batch 
corrected data.

```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
library(ggmap)
library(randomForest)
library(batchelor)
library(igraph)
library(cowplot)
library(ggpubr)
library(data.table)
```

Load spatial and segmentation functions.

```{r}
sapply(list.files("../functions/", full.names = TRUE), source)
```

# Load spatial cell and mRNA data

Output of TimEmbryos-E8.5.Rmd

```{r}
sce_filt = readRDS("../analysis_output/E8.5/E8.5_sce_filt_unlabelled.Rds")
sce_filt

cData_filt = as.data.frame(colData(sce_filt))
```

# Replace area normalised with library size normalised

```{r}
sce_filt <- reNormalise(sce_filt)
assay(sce_filt, "cosineNorm") <- cosineNorm(assay(sce_filt, "logcounts"))
```

# Load E8.25 and E8.5 atlas data

Also fix gene name for [Cavin3](http://www.informatics.jax.org/marker/MGI:1923422).

Also parse annotation of various celltypes.

```{r}
if (!file.exists("../analysis_output/E8.5/atlas_for_celltype_mapping.Rds")) {
  atlas = readRDS("../Data/atlas/sce_atlas_E8.25_E8.5.Rds")
  rowData(atlas)[rowData(atlas)[,"SYMBOL"] == "Prkcdbp", "SYMBOL"] <- "Cavin3"
  rowData(atlas)[rowData(atlas)[,"SymbolUniq"] == "Prkcdbp", "SymbolUniq"] <- "Cavin3"
  rownames(atlas) <- rowData(atlas)$SymbolUniq
  atlas <- computeSumFactors(atlas)
  atlas <- multiBatchNorm(atlas, batch = atlas$sample, preserve.single = TRUE)
  atlas
  head(colData(atlas))
  
  # Restrict atlas to cells that have a cell type label
  atlas = atlas[,!is.na(atlas$celltype)]
  atlas
  
  # Subset atlas down to E8.5
  atlas = atlas[, atlas$stage == "E8.5"]
  
  atlas
  
  # Batch aware normalise
  atlas <- multiBatchNorm(atlas, batch = atlas$sample, preserve.single = TRUE)
  saveRDS(atlas, file = "../analysis_output/E8.5/atlas_for_celltype_mapping.Rds")
} else {
  atlas = readRDS("../analysis_output/E8.5/atlas_for_celltype_mapping.Rds")
}

if (!file.exists("../analysis_output/E8.5/atlas_sub.Rds")) {
  
  # Restrict atlas to genes in SeqFISH
  atlas_sub = atlas[rowData(atlas)$SymbolUniq %in% rownames(sce_filt),]
  atlas_sub
  assay(atlas_sub, "cosineNorm") <- cosineNorm(logcounts(atlas_sub))
  
  ############################
  ############################
  ############################
  # Refine annotation
  
  # Collapse blood celltypes
  atlas_sub$celltype_parsed = atlas_sub$celltype
  atlas_sub$celltype_parsed[atlas_sub$celltype_parsed %in% c("Erythroid1", "Erythroid2", "Erythroid3")] <- "Erythroid"
  atlas_sub$celltype_parsed[atlas_sub$celltype_parsed %in% c("Blood progenitors 1", "Blood progenitors 2")] <- "Blood progenitors"
  
  # Expand paraxial and somitic mesoderm types
  paraxial = read.table("../Data/atlas/annotation/paraxialmeso_e85_annotation.txt",
                        header = TRUE, row.names = 1)
  table(paraxial$celltype, useNA = "always")
  table(atlas_sub[,rownames(paraxial)]$celltype, useNA = "always")
  
  paraxial_names = c("Head_mesoderm" = "Cranial mesoderm",
                     "Anterior-most_somites" = "Anterior somitic tissues",
                     "Sclerotome" = "Sclerotome",
                     "Dermomyotome" = "Dermomyotome")
  paraxial$celltypeparsed = paraxial_names[as.character(paraxial$celltype)]
  table(paraxial$celltype, paraxial$celltypeparsed)
  
  atlas_sub$celltype_parsed[match(rownames(paraxial), colnames(atlas_sub))] <- as.character(paraxial[,"celltypeparsed"])
  
  somitic = read.table("../Data/atlas/annotation/somiticmeso_e85_annotation.txt",
                       header = TRUE, row.names = 1)
  table(somitic$celltype, useNA = "always")
  table(atlas_sub[,rownames(somitic)]$celltype, useNA = "always")
  
  somitic_names = c("Posterior-most_somites" = "Posterior somitic tissues",
                    "Presomitic_mesoderm" = "Presomitic mesoderm")
  somitic$celltypeparsed = somitic_names[as.character(somitic$celltype)]
  table(somitic$celltype, somitic$celltypeparsed)
  
  atlas_sub$celltype_parsed[match(rownames(somitic), colnames(atlas_sub))] <- as.character(somitic[,"celltypeparsed"])
  
  # Posterior somites / pre-somitic mesoderm  only pre-somitic mesoderm
  # i.e. replace "Posterior somitic tissues" with "Presomitic mesoderm"
  atlas_sub$celltype_parsed[atlas_sub$celltype_parsed == "Posterior somitic tissues"] <- "Presomitic mesoderm"
  
  # Replace ExE mesoderm with "Lateral plate mesoderm"
  atlas_sub$celltype_parsed[atlas_sub$celltype_parsed == "ExE mesoderm"] <- "Lateral plate mesoderm"
  
  # Rename Gut to "Gut tube"
  atlas_sub$celltype_parsed[atlas_sub$celltype_parsed == "Gut"] <- "Gut tube"
  
  # Rename Def. endoderm to "Definitive endoderm"
  atlas_sub$celltype_parsed[atlas_sub$celltype_parsed == "Def. endoderm"] <- "Definitive endoderm"
  
  # Rename Pharyngeal mesoderm to "Splanchnic mesoderm"
  atlas_sub$celltype_parsed[atlas_sub$celltype_parsed == "Pharyngeal mesoderm"] <- "Splanchnic mesoderm"
  
  allCellTypes = sort(unique(atlas_sub$celltype_parsed))
  sort(table(atlas_sub$celltype_parsed))
  
  celltypesToKeep = names(which(table(atlas_sub$celltype_parsed) >= 25))
  
  celltypesToKeep
  print(setdiff(allCellTypes, celltypesToKeep))
  saveRDS(celltypesToKeep, file = "../analysis_output/E8.5/celltypesToKeep.Rds")
  
  atlas_sub <- atlas_sub[, atlas_sub$celltype_parsed %in% celltypesToKeep]
  saveRDS(atlas_sub, file = "../analysis_output/E8.5/atlas_sub.Rds")
} else {
  atlas_sub = readRDS("../analysis_output/E8.5/atlas_sub.Rds")
}

g = ggplot(cbind(as.data.frame(reducedDim(atlas_sub, "umap")),
                 celltype = atlas_sub$celltype),
           aes(x = x, y = y, colour = celltype)) + 
  geom_point() + 
  scale_colour_manual(values = celltype_colours) + 
  theme_classic() +
  NULL
g

g = ggplot(cbind(as.data.frame(reducedDim(atlas_sub, "umap")),
                 celltype = atlas_sub$celltype_parsed),
           aes(x = x, y = y, colour = celltype)) + 
  geom_point() + 
  scale_colour_manual(values = celltype_colours) + 
  theme_classic() +
  NULL
g
```

# Joint embedding of seqFISH and atlas data

Get the joint embedding for visualisation and initial classification. Also
identify the closest cell IDs within the joint embedding.

```{r}
joint_mnn_pca = getJointPCA(atlas_sub,
                            sce_filt,
                            assayNameAtlas = "logcounts",
                            assayNameSpatial = "logcounts",
                            spatial_batch = c("embryo","pos", "z"),
                            inc_z = TRUE,
                            irlba = FALSE,
                            multibatch_2 = FALSE)
saveRDS(joint_mnn_pca, "../analysis_output/E8.5/E8.5_joint_PCA_corrected_libsize.Rds")

joint_mnn_umap = calculateUMAP(joint_mnn_pca, transposed = TRUE)
rownames(joint_mnn_umap) <- rownames(joint_mnn_pca)
saveRDS(joint_mnn_umap, "../analysis_output/E8.5/E8.5_joint_UMAP_corrected_libsize.Rds")
```

# Extract KNN-based classification and quality score

```{r}
knns = BiocNeighbors::queryKNN(
  joint_mnn_pca[colnames(atlas_sub),],
  joint_mnn_pca[colnames(sce_filt),],
  k = 25, get.index = TRUE, get.distance = FALSE)

k.mapped = vectorSubset(colnames(atlas_sub), knns$index)
rownames(k.mapped) <- colnames(sce_filt)
saveRDS(k.mapped, "../analysis_output/E8.5/E8.5_naive_k.mapped_libsize.Rds")
predmat = vectorMatch(atlas_sub$celltype_parsed, k.mapped, colnames(atlas_sub))
knn_out = get_MVS_vals(predmat)
saveRDS(knn_out, "../analysis_output/E8.5/E8.5_naive_pred_knn_libsize.Rds")
```

# Get a joint PCA and metadata representation

```{r}
pred_knn_joint_unsorted = c(as.character(knn_out[,"celltype_mapped"]), as.character(atlas_sub$celltype_parsed))
names(pred_knn_joint_unsorted) <- c(rownames(knn_out), colnames(atlas_sub))

joint_df = data.frame(
  umap1 = joint_mnn_umap[,1],
  umap2 = joint_mnn_umap[,2],
  ctype = pred_knn_joint_unsorted[rownames(joint_mnn_umap)],
  type = ifelse(substring(rownames(joint_mnn_umap),1,1) == "e", "seqFISH", "atlas")
)
g0 = ggplot(
  joint_df, aes(x = umap1, y = umap2)) +
  geom_point(aes(colour = type), size = 1) +
  scale_colour_manual(values = c("seqFISH" = "red", "atlas" = "black")) +
  theme_classic() +
  theme(legend.position = "bottom") +
  # coord_fixed() +
  NULL
g0

g = ggplot(
  joint_df, aes(x = umap1, y = umap2)) +
  geom_point(aes(colour = ctype), size = 1) +
  scale_colour_manual(values = celltype_colours) +
  theme_classic() +
  theme(legend.position = "bottom") +
  NULL
g + facet_wrap(~type) + plot_layout(nrow = 1)
```

# Extract closest cells

```{r}
closest_atlas_cell_ind = BiocNeighbors::queryKNN(
  joint_mnn_pca[colnames(atlas_sub),],
  joint_mnn_pca[colnames(sce_filt),],
  k = 1, get.index = TRUE, get.distance = FALSE)
closest_atlas_cell = as.character(vectorSubset(colnames(atlas_sub), closest_atlas_cell_ind[[1]]))
names(closest_atlas_cell) <- colnames(sce_filt)
saveRDS(closest_atlas_cell, "../analysis_output/E8.5/E8.5_closest_atlas_cell_libsize.Rds")

closest_atlas_celltype = as.character(vectorMatch(
  atlas_sub$celltype_parsed, as.matrix(closest_atlas_cell),
  colnames(atlas_sub)))
names(closest_atlas_celltype) <- colnames(sce_filt)
saveRDS(closest_atlas_celltype, "../analysis_output/E8.5/E8.5_closest_atlas_celltype_libsize.Rds")

closest_spatial_cells_ind = do.call(cbind, lapply(split(colnames(sce_filt), interaction(sce_filt$embryo, sce_filt$z)), function(n) {
  n_ind = BiocNeighbors::queryKNN(
    joint_mnn_pca[n,],
    joint_mnn_pca[colnames(atlas_sub),],
    k = 1, get.index = TRUE, get.distance = FALSE)
  return(n[n_ind[[1]]])
}))
rownames(closest_spatial_cells_ind) <- colnames(atlas_sub)
saveRDS(closest_spatial_cells_ind, "../analysis_output/E8.5/E8.5_atlas_closest_spatial_cells_libsize.Rds")
```

# Jointly cluster and subcluster the atlas and spatial cells

```{r}
graph = buildSNNGraph(joint_mnn_pca, transposed = TRUE)
clusters_graph = cluster_louvain(graph)
clusters = as.numeric(membership(clusters_graph))
names(clusters) <- rownames(joint_mnn_pca)
table(clusters, useNA = "always")

subclusters = clusters
for (clust in unique(clusters)) {
  
  print(clust)
  
  # perform subclustering within this group
  graph = buildSNNGraph(joint_mnn_pca[clusters == clust,], transposed = TRUE)
  clusters_graph = cluster_louvain(graph)
  subcluster = as.numeric(membership(clusters_graph))
  names(subcluster) <- rownames(joint_mnn_pca)[clusters == clust]
  
  subclusters[clusters == clust] <- paste0(clust, ".", subcluster)
}
saveRDS(subclusters, "../analysis_output/E8.5/E8.5_joint_subclustering_libsize.Rds")
```

```{r}
tab = unclass(table(joint_df$ctype[joint_df$type == "atlas"], subclusters[joint_df$type == "atlas"]))

sort(apply(tab,2,function(x) sort(x, decreasing = TRUE)[2]))

sort(tab["Forebrain/Midbrain/Hindbrain",])

# for each subcluster, what is it's majority atlas class?
majority_subcluster = sapply(sort(unique(subclusters)), function(clust) {
  if (!clust %in% colnames(tab)) return(NA)
  names(sort(tab[,clust], decreasing = TRUE))[1]
})
table(majority_subcluster, useNA = "always")

knntab = unclass(table(joint_df$ctype[joint_df$type == "seqFISH"], subclusters[joint_df$type == "seqFISH"]))

majority_KNN = sapply(sort(unique(subclusters)), function(clust) {
  if (!clust %in% colnames(knntab)) return(NA)
  names(sort(knntab[,clust], decreasing = TRUE))[1]
})
table(majority_KNN, useNA = "always")


numCellsDF = cbind(data.frame(unclass(table(subclusters, joint_df$type))), majority_subcluster, majority_KNN)
numCellsDF

setdiff(as.character(joint_df$ctype), majority_subcluster)

joint_df$clust = subclusters

g = ggplot(
  joint_df, aes(x = umap1, y = umap2)) +
  geom_point(aes(colour = clust), size = 1) +
  theme_classic() +
  theme(legend.position = "bottom") +
  NULL
g + facet_wrap(~type) + plot_layout(nrow = 1)
```


```{r}
coord_x_named = c(sce_filt$x_global_affine, reducedDim(atlas_sub, "umap")[,1])
names(coord_x_named) <- c(colnames(sce_filt), colnames(atlas_sub))
coord_y_named = c(-sce_filt$y_global_affine, reducedDim(atlas_sub, "umap")[,2])
names(coord_y_named) <- c(colnames(sce_filt), colnames(atlas_sub))
split = c(paste0(sce_filt$embryo, sce_filt$z), rep("atlas", ncol(atlas_sub)))
names(split) <- c(colnames(sce_filt), colnames(atlas_sub))

joint_df$coord_x = coord_x_named[rownames(joint_df)]
joint_df$coord_y = coord_y_named[rownames(joint_df)]
joint_df$split = split[rownames(joint_df)]

brain_clusts = rownames(numCellsDF)[numCellsDF$majority_subcluster == "Forebrain/Midbrain/Hindbrain" & !is.na(numCellsDF$majority_subcluster)]

gList = sapply(brain_clusts, function(clustval) {
  g = ggplot(
    joint_df, aes(x = coord_x, y = coord_y)) +
    geom_point(aes(colour = clust == clustval), size = 0.2) +
    scale_colour_manual(values = c("TRUE" = "black", "FALSE" = "lightgrey")) +
    theme_classic() +
    theme(legend.position = "none") +
    ylim(clustval) +
    NULL
  return(g + facet_wrap(~split, scales = "free", nrow = 1))
}, simplify = FALSE)
g = wrap_plots(gList, ncol = 1)
ggsave(g, file = "../Figures/brain_joint_subclusters.png", height = 30, width = 20)
```

```{r}
sessionInfo()
```