---
title: "E8.5 seqFISH - Mesoderm analysis using Han et al Data"
author: "Shila Ghazanfar"
date: "23/10/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 16
                     fig_height: 12
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, cache.lazy = FALSE)
```

```{r}
set.seed(2020)
```

```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
library(ggmap)
library(randomForest)
library(batchelor)
library(igraph)
library(cowplot)
library(ggpubr)
library(Matrix)
library(grDevices)
library(RColorBrewer)
```

```{r}
sapply(list.files("../functions/", full.names = TRUE), source)
```

```{r}
na_colour = "white"
polygon_size = 0.1
polygon_colour = "black"
```

# Load Han data

Load Han et al processed data file

```{r}
processedFile_Han = "../Data/Han/processed/sce_han_all.Rds"
sce_han_raw = readRDS(processedFile_Han)
sce_han_raw
```

# Load Nowotschin data

```{r}
processedFile_Nowotschin = "../Data/Nowotschin/processed/Nowotschin_sce_sub.Rds"
sce_now_raw = readRDS(processedFile_Nowotschin)
sce_now_raw
```

# Load spatial data

```{r}
sce = readRDS("../analysis_output/E8.5/E8.5_sce_filt_unlabelled.Rds")
sce <- reNormalise(sce)
```

# Add cell type and clustering information

```{r}
mapping_dt = readRDS("../analysis_output/E8.5/celltype_annotation_refined.Rds")
colData(sce) <- cbind(colData(sce), mapping_dt[colnames(sce), setdiff(colnames(mapping_dt),colnames(colData(sce)))])
```

```{r}
meso_celltypes = c("Pharyngeal mesoderm",
                   "Splanchnic mesoderm",
                   "Paraxial mesoderm",
                   "Intermediate mesoderm",
                   "Mesenchyme",
                   "NMP",
                   "ExE mesoderm",
                   "Endothelium")
```

Subset to relevant cell types, using Tim's suggestion:
relevant mesodermal subtypes for Han data:
pharyngeal mesoderm / paraxialmesoderm /  intermediate mesoderm / Mesenchyme
Would also include: NMP / extraembryonic mesoderm / endothelium

# Subset and batch correction of Han data

Subset down to the E8.5, find out how many cell types they actually include, 
also figure out if we want to keep the other E8.75 samples. Batch correct them
using the sample information we have (need to pull from the cell names).

```{r}
head(colData(sce_han_raw))
table(sce_han_raw$stage)
table(sce_han_raw$cluster_annotated, sce_han_raw$stage, useNA = "always")
```

Restrict down to the E8.5 samples.

```{r}
han_samplesToKeep = as.character(unique(colData(sce_han_raw)[colData(sce_han_raw)$stage == "E8.5","cell"]))
head(han_samplesToKeep)

sce_han = sce_han_raw[,han_samplesToKeep]
sce_han
```

Restrict to genes that are in seqFISH data, fix gene names as needed.
[Cavin3](http://www.informatics.jax.org/marker/MGI:1923422)
[Nkx1-2](http://www.informatics.jax.org/marker/MGI:104806)
[Nkx2-3](http://www.informatics.jax.org/marker/MGI:97348)

```{r}
table(rownames(sce) %in% rownames(sce_han))
setdiff(rownames(sce), rownames(sce_han))

rownames(sce_han)[rownames(sce_han) == "Prkcdbp"] <- "Cavin3"

table(rownames(sce) %in% rownames(sce_han))
setdiff(rownames(sce), rownames(sce_han))
```

# Subset and batch correction of Nowotschin data

```{r}
head(colData(sce_now_raw))
sce_now_raw$Sample <- gsub("_.*", "", colnames(sce_now_raw))
table(sce_now_raw$Sample)
table(sce_now_raw$Sample, sce_now_raw$AP_pseudo, useNA = "always")

# Restrict down to the samples that contribute to the pseudospace.
now_samplesToKeep = unique(colData(sce_now_raw)[!is.na(sce_now_raw$AP_pseudo),"Sample"])
now_samplesToKeep
sce_now = sce_now_raw[, sce_now_raw$Sample %in% now_samplesToKeep]
sce_now
# Restrict to genes that are in seqFISH data, fix gene names as needed. Cavin3 Nkx1-2 Nkx2-3
table(rownames(sce) %in% rownames(sce_now))
setdiff(rownames(sce), rownames(sce_now))
rownames(sce_now)[rownames(sce_now) == "Prkcdbp"] <- "Cavin3"
rownames(sce_now)[rownames(sce_now) == "Nkx1.2"] <- "Nkx1-2"
rownames(sce_now)[rownames(sce_now) == "Nkx2.3"] <- "Nkx2-3"
table(rownames(sce) %in% rownames(sce_now))
setdiff(rownames(sce), rownames(sce_now))
```

# Joint mapping of Han and Nowotschin and spatial data

To avoid sex specific effects, remove Xist from the PCA calculation step.

```{r}
sce_han_sub <- sce_han[rownames(sce_han) %in% rownames(sce),][rownames(sce),]
sce_han_sub <- logNormCounts(sce_han_sub)

assay(sce_han_sub, "cosineNorm") <- cosineNorm(logcounts(sce_han_sub))
```

# Perform HVG gene selection of Han data

Only for those with labels

```{r}
sce_han <- logNormCounts(sce_han)
stats = modelGeneVar(sce_han[,!is.na(sce_han$cluster_annotated)], block = sce_han[,!is.na(sce_han$cluster_annotated)]$Sample)
hvgs_all = getTopHVGs(stats, fdr.threshold = 0.05) # more stringent
hvgs = intersect(hvgs_all, rownames(sce))
length(hvgs)
```

# Same for Nowotschin data

To avoid sex specific effects, remove Xist from the PCA calculation step.

```{r}
sce_now_sub <- sce_now[rownames(sce_now) %in% rownames(sce),][rownames(sce),]
sce_now_sub <- logNormCounts(sce_now_sub)

assay(sce_now_sub, "cosineNorm") <- cosineNorm(logcounts(sce_now_sub))
```

# Perform HVG gene selection of Han data

Only for those with labels

```{r}
sce_now <- logNormCounts(sce_now)
stats_now = modelGeneVar(sce_now[,!is.na(sce_now$AP_pseudo)], block = sce_now[,!is.na(sce_now$AP_pseudo)]$Sample)
hvgs_all_now = getTopHVGs(stats_now, fdr.threshold = 0.05) # more stringent
hvgs_now = intersect(hvgs_all_now, rownames(sce))
length(hvgs_now)
```

# concatenate the han and nowotschin data

```{r}
sce_han_now = SingleCellExperiment(
  assays = list(logcounts = cbind(assay(sce_han_sub, "logcounts"),
                                  assay(sce_now_sub, "logcounts")))
)
sce_han_now$stage = c(sce_han_sub$stage, sce_now_sub$Sample) 
table(sce_han_now$stage)
```

# get common hvgs

```{r}
length(hvgs)
length(hvgs_now)
hvgs
hvgs_now
hvgs_common = setdiff(union(hvgs, hvgs_now), "Xist")
length(hvgs_common)
hvgs_common
```

# Get joint PC space for Han and spatial data

```{r}
atlas = sce_han_now
spatial = sce
assayNameAtlas = "logcounts"
assayNameSpatial = "logcounts"
inc_z = TRUE
atlas_batch = "stage"
spatial_batch = c("embryo", "pos","z")
irlba = FALSE # if this is FALSE, then multibatchPCA

all_assay = cbind(assay(spatial, assayNameSpatial)[hvgs_common,],
                  # assay(atlas, assayNameAtlas)[rownames(spatial),])
                  assay(atlas, assayNameAtlas)[hvgs_common,])

if (!inc_z) {
  spatial_batch <- setdiff(spatial_batch, "z")
}

spatial_batchFactor = do.call(interaction, 
                              sapply(spatial_batch, function(x) 
                                colData(spatial)[,x], simplify = FALSE))

atlas_batchFactor = do.call(interaction, 
                            sapply(atlas_batch, function(x) 
                              colData(atlas)[,x], simplify = FALSE))

# make sure the levels of the atlas batchFactor are first
batchFactor = factor(c(as.character(spatial_batchFactor),
                       as.character(atlas_batchFactor)),
                     levels = c(sort(unique(as.character(atlas_batchFactor))),
                                sort(unique(as.character(spatial_batchFactor)))))
print(table(batchFactor))
levels(batchFactor)

# set the merge order
# we just have one group for the Han et al data

# MANUALLY CODED
# nowotschin clusters then han then so on..
merge.ordering = unlist(list(list(list(5,3,4,2,1)), as.list(6:length(levels(batchFactor)))), recursive = FALSE)

if (!file.exists("../analysis_output/E8.5/han_now_joint_mnn_pca.Rds")) {
  
  if (!irlba) {
    out = fastMNN(all_assay,
                  batch = batchFactor,
                  subset.row = !(rownames(all_assay) %in% "Xist"),
                  d = 50,
                  merge.order = merge.ordering)
    
    print(head(out@metadata$merge.info))
    
    joint_mnn_pca_han <- reducedDim(out, "corrected")
  } else {
    
    pca_all <- irlba::prcomp_irlba(t(all_assay), n = 50)$x
    rownames(pca_all) <- colnames(all_assay)
    
    out = reducedMNN(pca_all,
                     batch = batchFactor,
                     # subset.row = !(rownames(all_assay) %in% "Xist"),
                     # d = 50,
                     merge.order = merge.ordering)
    
    joint_mnn_pca_han <- out$corrected
    
  }
  
  saveRDS(joint_mnn_pca_han, file = "../analysis_output/E8.5/han_now_joint_mnn_pca.Rds")
} else {
  joint_mnn_pca_han = readRDS("../analysis_output/E8.5/han_now_joint_mnn_pca.Rds")
}

if (!file.exists("../analysis_output/E8.5/han_now_joint_mnn_umap.Rds")) {
  joint_mnn_umap_han = calculateUMAP(joint_mnn_pca_han, transposed = TRUE)
  rownames(joint_mnn_umap_han) <- rownames(joint_mnn_pca_han)
  saveRDS(joint_mnn_umap_han, file = "../analysis_output/E8.5/han_now_joint_mnn_umap.Rds")
} else {
  joint_mnn_umap_han = readRDS("../analysis_output/E8.5/han_now_joint_mnn_umap.Rds")
}

plot(joint_mnn_umap_han)


ds = substring(rownames(joint_mnn_umap_han),1,1)
ds[ds == "e"] <- "seqFISH"
ds[ds == "L"] <- "Nowotschin"
ds[ds %in% c("A", "C", "G","T")] <- "Han"
table(ds)

ind = sample(nrow(joint_mnn_umap_han))

plot(joint_mnn_umap_han[ind,], col = factor(ds)[ind], pch = 16, cex = 0.2)
legend("topright", col = 1:3, legend = levels(factor(ds)), pch = 16)

plot(joint_mnn_umap_han, col = factor(substring(rownames(joint_mnn_umap_han),1,1) %in% c("e", "L")), pch = 16, cex = 0.5)

ap = as.character(sce_han_sub$cluster_annotated)
names(ap) <- colnames(sce_han_sub)

table(sce_han_sub$celltype_broad, useNA = "always")
table(as.character(sce_han_sub$cluster_annotated), useNA = "always")

bc = as.character(sce_han_sub$celltype_broad)
names(bc) <- colnames(sce_han_sub)

ap_p = as.character(sce_now_sub$AP_pseudo)
names(ap_p) <- colnames(sce_now_sub)

plot(joint_mnn_umap_han, col = factor(ap[rownames(joint_mnn_umap_han)]),
     pch = 16, cex = 0.5)

plot(joint_mnn_umap_han, col = factor(bc[rownames(joint_mnn_umap_han)] == "Spl_Meso"),
     pch = 16, cex = 0.5)

plot(joint_mnn_umap_han, col = factor(ap_p[rownames(joint_mnn_umap_han)] == "Lung 1"),
     pch = 16, cex = 0.5)

joint_mnn_umap_han_spatial = joint_mnn_umap_han[c(colnames(sce), colnames(sce_han_sub)),]
```


# calculate nearest neighbour cells in four different ways

- seqFISH onto Han (for mesodermal subtypes)
- seqFISH onto Nowotschin (for endoderm subtypes consistency check)
- Han onto Nowotschin (for mesodermal subtypes)
- Nowotschin onto Han (for endoderm subtypes)

```{r}
# naming is query then reference
k.mapped_seqFISH_Han = get_k.mapped(joint_mnn_pca_han, colnames(sce), colnames(sce_han_sub))
k.mapped_seqFISH_Nowotschin = get_k.mapped(joint_mnn_pca_han, colnames(sce), colnames(sce_now_sub))
k.mapped_Han_Nowotschin = get_k.mapped(joint_mnn_pca_han, colnames(sce_han_sub), colnames(sce_now_sub))
k.mapped_Nowotschin_Han = get_k.mapped(joint_mnn_pca_han, colnames(sce_now_sub), colnames(sce_han_sub))
k.mapped_Nowotschin_seqFISH = get_k.mapped(joint_mnn_pca_han, colnames(sce_now_sub), colnames(sce))
k.mapped_Han_seqFISH = get_k.mapped(joint_mnn_pca_han, colnames(sce_han_sub), colnames(sce))

mapped_objs = intersect(paste0("k.mapped_", apply(expand.grid(c("Han", "Nowotschin", "seqFISH"),c("Han", "Nowotschin", "seqFISH")),1,paste0, collapse = "_")), ls())
mapped_objs
save(list = mapped_objs, file = "../analysis_output/E8.5/han_now_k.mapped_all.RData")

seqFISH_cluster_annotated = get_MVS_vals(vectorMatch(sce_han_sub$cluster_annotated, k.mapped_seqFISH_Han, colnames(sce_han_sub)))

seqFISH_AP_pseudo = get_MVS_vals(vectorMatch(sce_now_sub$AP_pseudo, k.mapped_seqFISH_Nowotschin, colnames(sce_now_sub)))

Han_AP_pseudo = get_MVS_vals(vectorMatch(sce_now_sub$AP_pseudo, k.mapped_Han_Nowotschin, colnames(sce_now_sub)))

Nowotschin_cluster_annotated = get_MVS_vals(vectorMatch(sce_han_sub$cluster_annotated, k.mapped_Nowotschin_Han, colnames(sce_han_sub)))

Nowotschin_celltype_mapped_refined = get_MVS_vals(vectorMatch(sce$celltype_mapped_refined, k.mapped_Nowotschin_seqFISH, colnames(sce)))

Han_celltype_mapped_refined = get_MVS_vals(vectorMatch(sce$celltype_mapped_refined, k.mapped_Han_seqFISH, colnames(sce)))

seqFISH_celltype_broad = get_MVS_vals(vectorMatch(sce_han_sub$celltype_broad, k.mapped_seqFISH_Han, colnames(sce_han_sub)))

Nowotschin_celltype_broad = get_MVS_vals(vectorMatch(sce_han_sub$celltype_broad, k.mapped_Nowotschin_Han, colnames(sce_han_sub)))

seqFISH_CellType = get_MVS_vals(vectorMatch(sce_now_sub$CellType, k.mapped_seqFISH_Nowotschin, colnames(sce_now_sub)))

Han_CellType = get_MVS_vals(vectorMatch(sce_now_sub$CellType, k.mapped_Han_Nowotschin, colnames(sce_now_sub)))


# seqfish, han, nowotschin
n_seqFISH = ncol(sce)
n_han = ncol(sce_han_sub)
n_now = ncol(sce_now_sub)
joint_df = data.frame(
  dataType = c(rep("seqFISH", n_seqFISH), rep("Han", n_han), rep("Nowotschin", n_now)),
  uniqueID = c(colnames(sce), colnames(sce_han_sub), colnames(sce_now_sub)),
  UMAP1 = c(joint_mnn_umap_han[colnames(sce),1],joint_mnn_umap_han[colnames(sce_han_sub),1],joint_mnn_umap_han[colnames(sce_now_sub),1]),
  UMAP2 = c(joint_mnn_umap_han[colnames(sce),2],joint_mnn_umap_han[colnames(sce_han_sub),2],joint_mnn_umap_han[colnames(sce_now_sub),2]),
  celltype_mapped_refined = c(as.character(sce$celltype_mapped_refined), rep(NA, n_han), rep(NA, n_now)),
  celltype_mapped_refined_all = c(as.character(sce$celltype_mapped_refined),
                                  Han_celltype_mapped_refined[,"celltype_mapped"],
                                  Nowotschin_celltype_mapped_refined[,"celltype_mapped"]),
  AP_pseudo = c(rep(NA, n_seqFISH), rep(NA, n_han), sce_now_sub$AP_pseudo),
  AP_pseudo_all = c(seqFISH_AP_pseudo[,"celltype_mapped"],
                    Han_AP_pseudo[,"celltype_mapped"],
                    sce_now_sub$AP_pseudo),
  cluster_annotated = c(rep(NA, n_seqFISH), as.character(sce_han_sub$cluster_annotated), rep(NA, n_now)),
  cluster_annotated_all = c(seqFISH_cluster_annotated[,"celltype_mapped"],
                            as.character(sce_han_sub$cluster_annotated),
                            Nowotschin_cluster_annotated[,"celltype_mapped"]),
  celltype_broad = c(rep(NA, n_seqFISH), as.character(sce_han_sub$celltype_broad), rep(NA, n_now)),
  celltype_broad_all = c(seqFISH_celltype_broad[,"celltype_mapped"],
                         as.character(sce_han_sub$celltype_broad),
                         Nowotschin_celltype_broad[,"celltype_mapped"]),
  CellType =  c(rep(NA, n_seqFISH), rep(NA, n_han), as.character(sce_now_sub$CellType)),
  CellType_all = c(seqFISH_CellType[,"celltype_mapped"],
                   Han_CellType[,"celltype_mapped"],
                   as.character(sce_now_sub$CellType))
)
```


```{r}
colnames(joint_df)

g = ggplot(joint_df[sample(nrow(joint_df)),],aes(x = UMAP1, y = UMAP2)) + 
  theme_classic() +
  coord_fixed() +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 5))) +
  NULL

g + geom_point(aes(colour = dataType), alpha = 0.3)

g + geom_point(aes(colour = celltype_mapped_refined), alpha = 0.3) + 
  scale_colour_manual(values = celltype_colours, na.value = "lightgrey")

g + geom_point(aes(colour = celltype_mapped_refined_all), alpha = 0.3) + 
  scale_colour_manual(values = celltype_colours, na.value = "lightgrey")

g + geom_point(aes(colour = AP_pseudo), alpha = 0.3) + 
  scale_colour_manual(values = AP_pseudo_colours, na.value = "lightgrey")

g + geom_point(aes(colour = AP_pseudo_all), alpha = 0.3) + 
  scale_colour_manual(values = AP_pseudo_colours, na.value = "lightgrey")

g + geom_point(aes(colour = cluster_annotated), alpha = 0.3)

g + geom_point(aes(colour = cluster_annotated_all), alpha = 0.3)

g + geom_point(aes(colour = celltype_broad), alpha = 0.3)

g + geom_point(aes(colour = celltype_broad_all), alpha = 0.3)

g + geom_point(aes(colour = CellType), alpha = 0.3)

g + geom_point(aes(colour = CellType_all), alpha = 0.3)
```

```{r}
saveRDS(joint_df, file = "../analysis_output/E8.5/E8.5_Han_Nowotschin_joint_mnn_df.Rds")
```

# Finish

```{r}
sessionInfo()
```