---
title: "E8.5 seqFISH - Combine E8.5 experiments and affine scaling"
author: "Shila Ghazanfar"
date: "04/06/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 10 
                     fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, cache.lazy = FALSE)
```

```{r}
set.seed(2020)
```

# Summary

In this report we take the individual experiment files for the E8.5 embryos, 
and output joint data files, with affine scaled coordinates. Note these 
objects contain 351 genes for the barcoded genes, not the extra 36 smFISH 
genes. 

```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
library(ggmap)
library(randomForest)
library(batchelor)
library(igraph)
library(cowplot)
library(ggpubr)
```

Load spatial and segmentation functions.

```{r}
library(SpatialUtils)
```

# Load mRNA data and merge

```{r}
mRNA_df_12 = readRDS("../Data/TimEmbryos-102219/processed/mRNA_df-102219.Rds")
head(mRNA_df_12)
mRNA_df_3 = readRDS("../Data/TimEmbryos-120919/processed/mRNA_df-120919.Rds")
head(mRNA_df_3)

mRNA_df = rbind(mRNA_df_12, mRNA_df_3)
rm(mRNA_df_12, mRNA_df_3)
```

# Load cell data and merge

```{r}
sce_12 = readRDS("../Data/TimEmbryos-102219/processed/sce_all-102219.Rds")
sce_12
sce_3 = readRDS("../Data/TimEmbryos-120919/processed/sce_all-120919.Rds")
sce_3

sum(counts(sce_12))
sum(counts(sce_3))

sce = cbind(sce_12,sce_3)
sce
colData(sce)
rm(sce_12, sce_3)
```

# Affine scaling of embryos in space

Affine scaling here consists of three steps:

* divide by the width/height of the fov square, here entered manually
* centre by subtracting the midpoint coordinate
* flip some embryos for the same top-down and left-right orientation
Having done this means that the embryos are on the same physical scale
to each other.

```{r}
# make a dataframe that says what needs to be done for the coordinates for each
# embryo
# 1. scale, 2. center, 3. flip
# still hardcoded so beware!

affine_df = data.frame(
  embryo = c("embryo1", "embryo2","embryo3"),
  scaleFactor_x = c(31,31,64),
  scaleFactor_y = c(31,31,64),
  midpoint_x = tapply(colData(sce)$x_global,colData(sce)$embryo,function(x)mean(range(x))),
  midpoint_y = tapply(colData(sce)$y_global,colData(sce)$embryo,function(x)mean(range(x))),
  flip_x = c(FALSE,TRUE,TRUE),
  flip_y = c(FALSE,FALSE,FALSE), 
  angle = c(0,-20,-35)
)

affine_df

angles = affine_df$angle
names(angles) <- rownames(affine_df)

x_global_scaled <- colData(sce)$x_global/affine_df[colData(sce)$embryo,"scaleFactor_x"]

y_global_scaled <- colData(sce)$y_global/affine_df[colData(sce)$embryo,"scaleFactor_y"]

x_global_scaled_centred = x_global_scaled - affine_df[colData(sce)$embryo,"midpoint_x"]/affine_df[colData(sce)$embryo,"scaleFactor_x"]

y_global_scaled_centred = y_global_scaled - affine_df[colData(sce)$embryo,"midpoint_y"]/affine_df[colData(sce)$embryo,"scaleFactor_y"]

x_global_scaled_centred_flipped = x_global_scaled_centred*ifelse(affine_df[colData(sce)$embryo, "flip_x"], -1, 1)

y_global_scaled_centred_flipped = y_global_scaled_centred*ifelse(affine_df[colData(sce)$embryo, "flip_y"], -1, 1)

df_global_scaled_centred_flipped = data.frame(embryo = sce$embryo,
                                              x = x_global_scaled_centred_flipped,
                                              y = y_global_scaled_centred_flipped)
out = rotateDF(df_global_scaled_centred_flipped, x = "x", y = "y", ang = angles)

colData(sce)$x_global_affine <- out$x
colData(sce)$y_global_affine <- out$y

# scaled and centred and flipped
g = ggplot(as.data.frame(colData(sce))) + 
  geom_point(aes(x = x_global_affine,
                 y = -y_global_affine,
                 colour = embryo), fill = NA, size = 0.5, alpha = 0.15) + 
  coord_fixed() +
  theme_classic() +
  ggtitle("All cells in affine scaling") +
  NULL
g

g + facet_grid(z~embryo)
```

Also do this with the segmentation numericLists. This takes a long time to run.

```{r}
x_global_segments <- colData(sce)$segmentation_vertices_x_global

x_global_segments_affine = sapply(names(x_global_segments), function(n) {
  
  # print(n)
  
  x_global_segment <- x_global_segments[[n]]
  if (length(x_global_segment) == 0) return(x_global_segment)
  
  embryo_val = colData(sce)[n,"embryo"]
  
  # 1. scale
  x_global_segment_scaled <- x_global_segment/affine_df[embryo_val,"scaleFactor_x"]
  
  # 2. center - note to scale the center too!
  x_global_segment_scaled_centred <- x_global_segment_scaled - affine_df[embryo_val,"midpoint_x"]/affine_df[embryo_val,"scaleFactor_x"]
  
  # 3. flip if needed
  if (affine_df[embryo_val,"flip_x"]) {
    x_global_segment_scaled_centred_flip = -x_global_segment_scaled_centred
  } else {
    x_global_segment_scaled_centred_flip <- x_global_segment_scaled_centred
  }
  
  return(x_global_segment_scaled_centred_flip)
}, simplify = FALSE)


y_global_segments <- colData(sce)$segmentation_vertices_y_global

y_global_segments_affine = sapply(names(y_global_segments), function(n) {
  
  # print(n)
  
  y_global_segment <- y_global_segments[[n]]
  if (length(y_global_segment) == 0) return(y_global_segment)
  
  embryo_val = colData(sce)[n,"embryo"]
  
  # 1. scale
  y_global_segment_scaled <- y_global_segment/affine_df[embryo_val,"scaleFactor_y"]
  
  # 2. center - note to scale the center too!
  y_global_segment_scaled_centred <- y_global_segment_scaled - affine_df[embryo_val,"midpoint_y"]/affine_df[embryo_val,"scaleFactor_y"]
  
  # 3. flip if needed
  if (affine_df[embryo_val,"flip_y"]) {
    y_global_segment_scaled_centred_flip = -y_global_segment_scaled_centred
  } else {
    y_global_segment_scaled_centred_flip <- y_global_segment_scaled_centred
  }
  
  return(y_global_segment_scaled_centred_flip)
}, simplify = FALSE)

x_global_segments_affine_rotated = x_global_segments_affine
y_global_segments_affine_rotated = y_global_segments_affine

# a list of dataframes

for (n in names(y_global_segments)) {
  
  # print(n)
  
  if (length(x_global_segments_affine[[n]]) == 0) next
  
  if (FALSE) {
    plot(x_global_segments_affine[[n]],
         y_global_segments_affine[[n]], type = "l", xlim = c(-3,3), ylim= c(-3,3))
  }
  
  df_global_scaled_centred_flipped = data.frame(embryo = as.character(colData(sce)[n,"embryo"]),
                                                x = x_global_segments_affine[[n]],
                                                y = y_global_segments_affine[[n]])
  out = rotateDF(df_global_scaled_centred_flipped, xname = "x", yname = "y", ang = angles)
  
  if (FALSE) {
    points(out$x,
         out$y, type = "l", col = "blue")
  }
  
  x_global_segments_affine_rotated[[n]] <- out$x
  y_global_segments_affine_rotated[[n]] <- out$y
  
}

colData(sce)$segmentation_vertices_x_global_affine <- IRanges::NumericList(x_global_segments_affine_rotated)
colData(sce)$segmentation_vertices_y_global_affine <- IRanges::NumericList(y_global_segments_affine_rotated)
```

Also do this for the mRNA dataframes

```{r}
# 1. scale, 2. center, 3. flip if needed
mRNA_x_global_affine <- mRNA_df$x_global
for (embryo_val in as.character(unique(affine_df$embryo))) {
  mRNA_x_global_affine[mRNA_df$embryo == embryo_val] <-
    ((mRNA_df$x_global[mRNA_df$embryo == embryo_val]/affine_df[embryo_val,"scaleFactor_x"]) - affine_df[embryo_val,"midpoint_x"]/affine_df[embryo_val,"scaleFactor_x"])*(ifelse(affine_df[embryo_val,"flip_x"], -1, 1))
}

mRNA_y_global_affine <- mRNA_df$y_global
for (embryo_val in as.character(unique(affine_df$embryo))) {
  mRNA_y_global_affine[mRNA_df$embryo == embryo_val] <-
    ((mRNA_df$y_global[mRNA_df$embryo == embryo_val]/affine_df[embryo_val,"scaleFactor_y"]) - affine_df[embryo_val,"midpoint_y"]/affine_df[embryo_val,"scaleFactor_y"])*(ifelse(affine_df[embryo_val,"flip_y"], -1, 1))
}

df_global_scaled_centred_flipped = data.frame(embryo = mRNA_df$embryo,
                                              x = mRNA_x_global_affine,
                                              y = mRNA_y_global_affine)
out = rotateDF(df_global_scaled_centred_flipped, x = "x", y = "y", ang = angles)

mRNA_df$x_global_affine <- out$x
mRNA_df$y_global_affine <- out$y
```

# Graph using boundaries

```{r}
boundary_polygons = getSegmentationVerticesDF(
  colData(sce),
  xname = "segmentation_vertices_x_global_affine",
  yname = "segmentation_vertices_y_global_affine",
  othercols = c("uniqueID","z","embryo","pos"))

boundary_bbox = do.call(
  rbind,
  lapply(
    split.data.frame(boundary_polygons, 
                     list(boundary_polygons$pos,
                          boundary_polygons$embryo,
                          boundary_polygons$z)),
    function(df) cbind(as.data.frame(t(make_bbox(segmentation_vertices_x_global_affine,
                                                 segmentation_vertices_y_global_affine,
                                                 data = df, f = 0))),
                       embryo = df$embryo[1],
                       pos = df$pos[1],
                       z = df$z[1]))
)
boundary_bbox <- na.omit(boundary_bbox)
```


```{r, fig.height=24}
for (zval in unique(boundary_polygons$z)) {
  
  g = ggplot(subset(as.data.frame(colData(sce)), z %in% zval)) + 
    geom_point(aes(group = uniqueID,
                     x = x_global_affine,
                     y = -y_global_affine), fill = NA, colour = "black") + 
    facet_wrap(embryo~., nrow = 3) +
    coord_fixed() +
    theme_classic() +
    ggtitle(zval) +
    NULL
  print(g)
}
```

```{r, fig.height=24}
for (zval in unique(boundary_polygons$z)) {
  
  g = ggplot(subset(boundary_polygons, z %in% zval)) + 
    geom_polygon(aes(group = uniqueID,
                     x = segmentation_vertices_x_global_affine,
                     y = -segmentation_vertices_y_global_affine), fill = NA, colour = "black") + 
    facet_wrap(embryo~., nrow = 3) +
    coord_fixed() +
    theme_classic() +
    theme(legend.position = "none") +
    ggtitle(zval) +
    NULL
  print(g)
  
  gg = g + geom_rect(data = subset(boundary_bbox, z %in% zval),
                     aes(xmin = left, 
                         xmax = right,
                         ymin = -top,
                         ymax = -bottom),
                     color="red", fill= "transparent", linetype = "dotted") +
    ggtitle("Horizontal Bounding box added") +
    NULL    
  print(gg)
}
```

# Basic Exploration

Calculate QC metrics and extract the column data as a separate object

```{r}
sce <- scater::addPerCellQC(sce)
cData = as.data.frame(colData(sce))
```

```{r}
cData$embryo_pos = factor(paste0(cData$embryo,"_", cData$pos),
                          levels = mixedsort(unique(as.character(paste0(cData$embryo,"_", cData$pos)))))
sce$embryo_pos <- cData$embryo_pos
mRNA_df$embryo_pos = factor(paste0(mRNA_df$embryo,"_", mRNA_df$pos),
                            levels = mixedsort(unique(as.character(paste0(mRNA_df$embryo,"_", mRNA_df$pos)))))


cData$embryo_pos_z = factor(paste0(cData$embryo,"_", cData$pos, "_", cData$z),
                          levels = mixedsort(unique(as.character(paste0(cData$embryo,"_", cData$pos, "_", cData$z)))))
sce$embryo_pos_z <- cData$embryo_pos_z
mRNA_df$embryo_pos_z = factor(paste0(mRNA_df$embryo,"_", mRNA_df$pos, "_", mRNA_df$z),
                            levels = mixedsort(unique(as.character(paste0(mRNA_df$embryo,"_", mRNA_df$pos, "_", mRNA_df$z)))))
```

# Save output

```{r}
sce
head(colData(sce))
dim(mRNA_df)
head(mRNA_df)
```

```{r}
saveRDS(sce, file = "../analysis_output/E8.5/E8.5_sce_unfiltered.Rds")
saveRDS(mRNA_df, file = "../analysis_output/E8.5/E8.5_mRNA_df_unfiltered.Rds")
```

```{r}
sessionInfo()
```