---
title: "Generate shiny input files"
author: "Shila Ghazanfar"
date: "01/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
library(ggmap)
library(randomForest)
library(batchelor)
library(igraph)
library(cowplot)
library(ggpubr)
library(Matrix)
library(grDevices)
library(reshape)
```

# Summary

The idea here is to generate the input files necessary for the shiny app, 
to avoid having to unnecessarily repeat various functions, and so that we do not
require a latest version of SingleCellExperiment

# Load data objects

```{r}
# load the functions
library(SpatialUtils)

# celltype colours
source("celltype_colours.R")

# load the data
sce = readRDS("../analysis_output/E8.5/E8.5_sce_filt_unlabelled.Rds")
sce <- reNormalise(sce)

mapping_dt = readRDS("../analysis_output/E8.5/celltype_annotation_refined.Rds")
colData(sce) <- cbind(colData(sce), mapping_dt[colnames(sce), setdiff(colnames(mapping_dt),colnames(colData(sce)))])
```

# Load fastMNN and quantile normalise expression data per embryo and z-stack

```{r}
if (!file.exists( "../analysis_output/E8.5/exprs_norm.Rds")) {
    mnn_out = readRDS("../analysis_output/E8.5/E8.5_fastMNN.Rds")
    mnn_out
    assay(sce, "fastMNN") <- assay(mnn_out, "reconstructed")
    
    sce$embryo_z <- interaction(sce$embryo, sce$z)
    
    sce_split = splitSCE(sce, sce$embryo_z)
    
    exprs_normList <- lapply(sce_split, function(sce_sub) {
        mnn = assay(sce_sub, "fastMNN")
        lcounts = assay(sce_sub, "logcounts")
        
        mnn_quantile = mnn
        
        for (i in 1:nrow(mnn_quantile)) {
            
            vals = mnn_quantile[i,]
            vals[order(vals)] <- sort(lcounts[i,])
            vals[lcounts[i,] == 0] <- 0
            mnn_quantile[i,] <- vals
            
        }
        
        return(mnn_quantile)
        
    })
    
    exprs_norm <- do.call(cbind, exprs_normList)[,colnames(sce)]
    exprs_norm <- as.matrix(exprs_norm)
    saveRDS(exprs_norm, file = "../analysis_output/E8.5/exprs_norm.Rds")
} else {
    exprs_norm = readRDS("../analysis_output/E8.5/exprs_norm.Rds")
}
```

# Extract metadata

```{r}
meta = as.data.frame(colData(sce))
meta <- updateMesenchymeLabel(meta)

celltypes = sort(as.character(unique(meta$celltype_mapped_refined)))
```

# Generate boundary polygons

```{r}
embryos = sort(as.character(unique(meta$embryo)))

zvals = sort(unique(meta$z))

genes = rownames(sce)

metanames = colnames(meta)

boundary_polygons = getSegmentationVerticesDF(
    as(meta, "DataFrame"),
    xname = "segmentation_vertices_x_global_affine",
    yname = "segmentation_vertices_y_global_affine",
    othercols = c("uniqueID",
                  "z","embryo",
                  "celltype_mapped_refined"))
rownames(boundary_polygons) <- NULL

boundary_polygons$segmentation_vertices_y_global_affine_neg <- -boundary_polygons$segmentation_vertices_y_global_affine

boundary_polygons <- boundary_polygons[,c("uniqueID",
                                          "z","embryo",
                                          "celltype_mapped_refined",
                                          "segmentation_vertices_x_global_affine",
                                          "segmentation_vertices_y_global_affine_neg")]

boundary_polygons$uniquePoint <- !duplicated(boundary_polygons$uniqueID)
```

# Split SingleCellExperiment into metadata and expression

```{r}
exprs = assay(sce, "logcounts")

dim(exprs)
exprs[1:5,1:5]
length(celltypes)
head(celltypes)
dim(meta)
head(meta)
dim(boundary_polygons)
head(boundary_polygons)
```

Convert h5 imputed matrix to lower precision, to reduce filesize for shinyapps
bundle size restriction

```{r}
library(HDF5Array)
imp <- HDF5Array(filepath = "../analysis_output/imputation/combined.h5", name = "logcounts")
if (!file.exists("../analysis_output/imputation/combined_compressed.h5")) {
    writeHDF5Array(imp, filepath="../analysis_output/imputation/combined_compressed.h5", name="logcounts", H5type="H5T_IEEE_F32LE", level=9)
}
```

# generate legend ggplots for app

```{r}

g_leg_list = list()

########## 

for (ttl in c("Imputed gene expression", "Batch-corrected gene expression")) {
    
    # get the legend
    pc_df = data.frame(Expression = 1:10)
    
    g_leg_raw = ggplot(pc_df, aes(x = Expression, y = Expression)) + 
        geom_point(aes(colour = Expression)) + 
        theme_transparent() + 
        scale_colour_gradientn(
            colours = c("gray75","cornflowerblue", "black"),
            labels = c("Low", "High"), 
            breaks = range(pc_df$Expression),
            limits = range(pc_df$Expression)) + 
        theme(legend.position = "bottom") +
        theme(legend.text = element_text(size = 10),
              legend.title = element_text(size = 15)) +
        guides(color = guide_colourbar(title.position = "top",
                                       title.hjust = 0.5,
                                       ticks = FALSE,
                                       title = ttl,
                                       barwidth = unit(1.5, "strwidth",
                                                       data = ttl
                                       ))) +
        NULL
    
    g_leg = as_ggplot(get_legend(g_leg_raw))
    g_leg_list[[ttl]] <- g_leg
    
}


##########

ttl = "Mapped cell type"

# get the legend
pc_df = data.frame(Expression = meta$celltype_mapped_refined)

g_leg_raw = ggplot(pc_df, aes(x = Expression, y = Expression)) +
    geom_point(aes(colour = Expression), size = 5) +
    theme_transparent() +
    scale_colour_manual(values = celltype_colours) +
    theme(legend.position = "bottom") +
    theme(legend.text = element_text(size = 15),
          legend.title = element_text(size = 20)) +
    guides(color = guide_legend(title.position = "top",
                                title.hjust = 0.5,
                                ticks = FALSE,
                                title = ttl)) +
    NULL

g_leg = as_ggplot(get_legend(g_leg_raw))
g_leg_list[[ttl]] <- g_leg
```


# Output files

The R objects generated go into the apps/data folder locally. We have multiple 
objects, a genes x cells logcounts matrix (sparse matrix), a cells x features 
metadata object (data.frame), and a vertices x features boundary polygons 
object (data.frame), celltypes (character vector), and imputed gene expression
(h5 file and two character vectors files).

```{r}
appDir = "../apps/SpatialMouseAtlas/"

if (!file.exists(paste0(appDir, "data"))) {
    system(paste0("mkdir ", appDir, "data"))
}

saveRDS(g_leg_list, file = paste0(appDir, "data/g_leg_list.Rds"))
saveRDS(as(exprs, "sparseMatrix"), file = paste0(appDir, "data/exprs.Rds"))
saveRDS(as(exprs_norm, "sparseMatrix"), file = paste0(appDir, "data/exprs_norm.Rds"))
saveRDS(meta, file = paste0(appDir, "data/meta.Rds"))
saveRDS(genes, file = paste0(appDir, "data/genes.Rds"))
saveRDS(boundary_polygons, file = paste0(appDir, "data/boundary_polygons.Rds"))
saveRDS(celltypes, file = paste0(appDir, "data/celltypes.Rds"))
```

# Take subset of mRNA df

```{r}
mRNA_df = readRDS("../analysis_output/E8.5/E8.5_mRNA_df_unfiltered.Rds")
mRNA_df <- mRNA_df[,c("geneID","z","embryo","x_global_affine","y_global_affine")]
saveRDS(mRNA_df, file = paste0(appDir, "data/mRNA.Rds"))
```

# Misc

Copy celltype_colours.R to the app directory.

And copy imputed to app directory.

```{r}
if (!file.exists(paste0(appDir, "celltype_colours.R"))) {
    system(paste0("cp celltype_colours.R ", appDir))
}
if (!file.exists(paste0(appDir, "/data/combined_compressed.h5"))) {
    system(paste0("cp ../analysis_output/imputation/combined_compressed.h5 ", appDir, "data"))
}
system(paste0("cp ../analysis_output/imputation/combined_cnames.Rds ", appDir, "data"))
system(paste0("cp ../analysis_output/imputation/combined_rnames.Rds ", appDir, "data"))

```

# Save files into content server folder

```{r}
meta_cols = c("uniqueID", "embryo", "pos", "z", "x_global", "y_global", "x_global_affine", "y_global_affine", "embryo_pos", "embryo_pos_z", "Area", "UMAP1", "UMAP2", "celltype_mapped_refined", "segmentation_vertices_x_global_affine", "segmentation_vertices_y_global_affine")

meta2 = meta[,meta_cols]

saveRDS(meta2, file = "../content/metadata.Rds")

counts = assay(sce, "counts")

saveRDS(as(exprs, "sparseMatrix"), file = "../content/exprs.Rds")
saveRDS(as(counts, "sparseMatrix"), file = "../content/counts.Rds")

# mRNA
mRNA_df = readRDS("../analysis_output/E8.5/E8.5_mRNA_df_unfiltered.Rds")
mRNA_df_content <- mRNA_df[,c("uniqueID", "geneID", "pos", "x", "y", "x_global", "y_global", "x_global_affine", "y_global_affine", "z", "seeds", "intensity")]
saveRDS(mRNA_df_content, file = "../content/mRNA.Rds")

boundary_polygons = getSegmentationVerticesDF(
    as(meta, "DataFrame"),
    xname = "segmentation_vertices_x_global_affine",
    yname = "segmentation_vertices_y_global_affine",
    othercols = c("uniqueID",
                  "z","embryo"))
rownames(boundary_polygons) <- NULL

saveRDS(boundary_polygons, file = "../content/segmentation_vertices.Rds")

if (!file.exists("../content/combined_compressed.h5")) {
    system(paste0("cp ../analysis_output/imputation/combined.h5 ", "../content/imputed.h5"))
}
system(paste0("cp ../analysis_output/imputation/combined_cnames.Rds ", "../content/imputed_column_names.Rds"))
system(paste0("cp ../analysis_output/imputation/combined_rnames.Rds ", "../content/imputed_row_names.Rds"))
```

# Finish

```{r}
sessionInfo()
```