---
title: "E8.5 seqFISH - Processing and exploratory analysis"
author: "Shila Ghazanfar"
date: "05/06/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 10 
                     fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE, cache.lazy = FALSE)
```

```{r}
set.seed(2020)
```

# Summary

In this report we perform joint analysis of three E8.5 embryos, and output
joint data files with associated measured and predicted metadata. Note these 
objects contain 351 genes for the barcoded genes, not the extra 36 smFISH genes. 

```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
library(ggmap)
library(randomForest)
library(batchelor)
library(igraph)
library(cowplot)
library(ggpubr)
```

Load spatial and segmentation functions.

```{r}
source("spatial_functions.R")
source("segmentation_functions.R")
source("celltype_colours.R")
source("neighbourSegments.R")
```

# Load spatial cell and mRNA data

```{r}
sce = readRDS("../analysis_output/E8.5/E8.5_sce_unfiltered.Rds")
mRNA_df = readRDS("../analysis_output/E8.5/E8.5_mRNA_df_unfiltered.Rds")
```

# Graph using boundaries

```{r}
boundary_polygons = getSegmentationVerticesDF(
  colData(sce),
  xname = "segmentation_vertices_x_global_affine",
  yname = "segmentation_vertices_y_global_affine",
  othercols = c("uniqueID","z","embryo","pos"))

boundary_bbox = do.call(
  rbind,
  lapply(
    split.data.frame(boundary_polygons, 
                     list(boundary_polygons$pos,
                          boundary_polygons$embryo,
                          boundary_polygons$z)),
    function(df) cbind(as.data.frame(t(make_bbox(segmentation_vertices_x_global_affine,
                                                 segmentation_vertices_y_global_affine,
                                                 data = df, f = 0))),
                       embryo = df$embryo[1],
                       pos = df$pos[1],
                       z = df$z[1]))
)
boundary_bbox <- na.omit(boundary_bbox)
```

```{r, fig.height=24}
for (zval in unique(boundary_polygons$z)) {
  
  g = ggplot(subset(boundary_polygons, z %in% zval)) + 
    geom_polygon(aes(group = uniqueID,
                     x = segmentation_vertices_x_global_affine,
                     y = -segmentation_vertices_y_global_affine), fill = NA, colour = "black") + 
    facet_wrap(embryo~z, nrow = 3) +
    coord_fixed() +
    theme_classic() +
    NULL
  print(g)
  
  gg = g + geom_rect(data = subset(boundary_bbox, z %in% zval),
                     aes(xmin = left, 
                         xmax = right,
                         ymin = -top,
                         ymax = -bottom),
                     color="red", fill= "transparent") + 
    NULL
  print(gg)
}
```

# Basic Exploration

Calculate QC metrics and extract the column data as a separate object

```{r}
cData = as.data.frame(colData(sce))
```

```{r}
ggplot(cData,
       aes(x = embryo_pos, y = log10(1+total),
           fill = embryo)) +
  geom_boxplot(outlier.size = 0.5) + 
  ggtitle("Total counts per FOV") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "bottom") +
  NULL

ggplot(cData,
       aes(x = embryo_pos, y = log10(1+total),
           fill = embryo)) +
  geom_boxplot(outlier.size = 0.5) + 
  ggtitle("Total counts per FOV") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "bottom") +
  facet_wrap(~z) +
  NULL

ggplot(cData,
       aes(x = embryo, y = log10(1+total))) +
  geom_boxplot(outlier.size = 0.5) + 
  ggtitle("Total counts per cell") +
  theme_classic() +
  NULL

ggplot(cData,
       aes(x = embryo, y = log10(1+total))) +
  geom_boxplot(outlier.size = 0.5) + 
  ggtitle("Total counts per cell") +
  theme_classic() +
  facet_wrap(~z) +
  NULL

ggplot(cData,
       aes(x = embryo, y = detected)) +
  geom_boxplot(outlier.size = 0.5) + 
  ggtitle("Number of genes detected per cell") +
  theme_classic() +
  NULL

ggplot(cData,
       aes(x = embryo, y = detected)) +
  geom_boxplot(outlier.size = 0.5) + 
  ggtitle("Number of genes detected per cell") +
  theme_classic() +
  facet_wrap(~z) +
  NULL

ggplot(cData,
       aes(x = embryo_pos, y = detected,
           fill = embryo)) +
  geom_boxplot(outlier.size = 0.5) + 
  ggtitle("Number of genes detected per FOV") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  NULL

ggplot(cData,
       aes(x = embryo_pos, y = detected,
           fill = embryo)) +
  geom_boxplot(outlier.size = 0.5) + 
  ggtitle("Number of genes detected per FOV") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~z) +
  NULL


ggplot(cData,
       aes(x = embryo, y = detected)) +
  geom_boxplot(outlier.size = 0.5) + 
  ggtitle("Number of genes detected per cell") +
  theme_classic() +
  NULL

ggplot(cData,
       aes(x = embryo, y = detected)) +
  geom_boxplot(outlier.size = 0.5) + 
  ggtitle("Number of genes detected per cell") +
  theme_classic() +
  facet_wrap(~z) +
  NULL
```

Is there a boundary effect?

```{r}
mid = function(x) {
  mean(range(x))
}
cData$x_dist = abs(unsplit(tapply(cData$x_global_affine, cData$embryo_pos, mid), cData$embryo_pos) - cData$x_global_affine) 
cData$y_dist = abs(unsplit(tapply(cData$y_global_affine, cData$embryo_pos, mid), cData$embryo_pos) - cData$y_global_affine) 
cData$max_dist = pmax(cData$x_dist, cData$y_dist)
cData$distanceToCentre = unsplit(tapply(cData$max_dist, cData$embryo_pos, function(x)x/max(abs(x))), cData$embryo_pos)

ggplot(cData,
       aes(x = x_global_affine, y = -y_global_affine, colour = distanceToCentre)) + 
  geom_point(size = 0.2) + 
  theme_classic() + 
  coord_fixed() +
  scale_colour_gradient(low = "black", high = "red") +
  facet_wrap(~embryo) +
  NULL

ggplot(cData,
       aes(x = distanceToCentre, y = detected,
           colour = embryo)) +
  geom_point(alpha = 0.2) + 
  ggtitle("Total genes detected per cell") +
  facet_wrap(~embryo) +
  theme_classic() +
  NULL
cor(cData$distanceToCentre, cData$detected)

ggplot(cData,
       aes(x = distanceToCentre^4, y = detected,
           colour = embryo)) +
  geom_point(alpha = 0.2) + 
  ggtitle("Total genes detected per cell") +
  facet_wrap(~embryo) +
  theme_classic() +
  NULL

ggplot(cData,
       aes(x = cut(distanceToCentre, 50), y = detected,
           colour = embryo,
           fill = embryo)) +
  geom_boxplot(outlier.size = 0.5) + 
  facet_wrap(embryo~z) +
  ggtitle("Total genes detected per cell") +
  theme(axis.text.x = element_text(angle = 90)) +
  theme_classic() +
  NULL

ggplot(cData,
       aes(x = cut(distanceToCentre, 50), y = log10(1+total),
           colour = embryo,
           fill = embryo)) +
  geom_boxplot() + 
  facet_wrap(embryo~z) +
  ggtitle("Total counts per cell") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  NULL
```

We observe a very slight negative correlation between distance to the centre
of the FOV and the expression detected, this really seems to be an issue right
on the border of the FOV square.

# Identify outlier segments that are unlikely to be cells

```{r}
ggplot(cData,
       aes(x = log10(Area),
           colour = embryo)) +
  geom_density() +
  facet_wrap(~z) +
  ggtitle("Cell area") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  NULL

ggplot(cData,
       aes(x = log10(Area),
           colour = factor(z))) +
  geom_density() +
  facet_wrap(~embryo) +
  ggtitle("Cell area") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  NULL

ggplot(cData,
       aes(x = log10(Area),
           colour = embryo_pos)) +
  geom_density() +
  facet_wrap(~z) +
  ggtitle("Cell area split by field of views") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none") +
  NULL

ggplot(cData,
       aes(x = log10(Area),
           group = interaction(embryo_pos, z),
           colour = factor(z))) +
  geom_density(alpha = 0.5) +
  ggtitle("Cell area split by field of views") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none") +
  NULL

ggplot(cData,
       aes(x = log10(Area), y = detected,
           colour = embryo,
           alpha = 0.1)) +
  geom_point() +
  facet_wrap(~z) +
  ggtitle("Total genes detected per FOV") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  NULL

ggplot(cData,
       aes(x = log10(Area), y = detected,
           colour = factor(z),
           alpha = 0.1)) +
  geom_point() +
  facet_wrap(~embryo) +
  ggtitle("Total genes detected per FOV") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  NULL

ggplot(cData,
       aes(x = log10(Area), y = log10(1+total),
           colour = embryo,
           alpha = 0.1)) +
  geom_point() +
  facet_wrap(~z) +
  ggtitle("Total counts detected per FOV") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  NULL

ggplot(cData,
       aes(x = log10(Area), y = log10(1+total),
           colour = factor(z),
           alpha = 0.1)) +
  geom_point() +
  facet_wrap(~embryo) +
  ggtitle("Total counts detected per FOV") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  NULL
```

There are a subset of "cells" with unrealistically large size. We pose that
these are segmentation objects that do not correspond to cells and should be 
removed.

We identify outlying "cells" by assuming that each cell is roughly spherical,
and that its cross-sectional diameter is roughly normally distributed. We also
note that the distributions across embryos and across positions are highly 
similar, so we can perform a single test for all cells. At a FDR-corrected
P-value level of 0.01, we test the (one-sided) deviation from normality for 
cells.

```{r}
filterAreas

cData$outlierSize = !filterAreas(cData$Area)

table(cData$outlierSize)

ggplot(cData,
       aes(x = log10(Area), y = detected,
           colour = outlierSize,
           alpha = 0.1)) +
  geom_point() +
  facet_grid(embryo~z) +
  ggtitle("Total genes detected") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  NULL

ggplot(cData,
       aes(x = log10(Area), y = log10(1+total),
           colour = outlierSize,
           alpha = 0.1)) +
  geom_point() +
  facet_grid(embryo~z) +
  ggtitle("Total counts detected") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) +
  NULL
```

# Filter for high quality cells

We remove cells according to three criteria:
* too large area
* too few genes
* too low counts

```{r}
keepCriteriaMatrix = 
  cbind(
    area_outlier = (!cData$outlierSize),
    genes_detected = (cData$detected >= 5),
    genes_total = (cData$total >= 10)
  )
head(keepCriteriaMatrix)

UpSetR::upset(as.data.frame(1*keepCriteriaMatrix))

UpSetR::upset(as.data.frame(1*keepCriteriaMatrix),
              scale.intersections = "log10")

keepCriteria <- apply(keepCriteriaMatrix,1,all)

table(keepCriteria)

sce_filt = sce[,keepCriteria]
cData_filt = cData[keepCriteria,]

mRNA_df_filt <- subset(mRNA_df, uniqueID %in% colnames(sce_filt))

boundary_polygons <- subset(boundary_polygons, uniqueID %in% sce_filt$uniqueID)

sce_filt
dim(mRNA_df_filt)
```

# Normalise counts to cell area

Assign the per-cell size factor as the square root of the area divided by the 
mean area across all cells. Also perform typical library size scaling, for 
potential comparisons in future.

```{r}
libsize_scaling = edgeR::calcNormFactors(counts(sce_filt))
sizeFactors(sce_filt) <- libsize_scaling
assay(sce_filt, "logcounts_libsize") <- logcounts(scater::logNormCounts(sce_filt))

out_area = getAreaFactorsDirectly(sce_filt, transform = NULL)
assay(sce_filt, "logcounts_area") <- out_area[["logcounts.area"]]

# use area norm function defined in spatial_functions.R
out_sqrt_area = getAreaFactorsDirectly(sce_filt, transform = sqrt)
assay(sce_filt, "logcounts_sqrtarea") <- out_sqrt_area[["logcounts.area"]]
# sizeFactors(sce_filt) <- out_area[["sizeFactors"]]
sizeFactors(sce_filt) <- NULL
```

```{r}
par(mfrow = c(2,2))
plot(colSums(counts(sce_filt)), sce_filt$Area, col = alpha("black", 0.3))
plot(colSums(counts(sce_filt)), sqrt(sce_filt$Area), col = alpha("black", 0.3))
plot(colSums(counts(sce_filt)), log10(sce_filt$Area), col = alpha("black", 0.3))
plot(colSums(counts(sce_filt)), 1/libsize_scaling, col = alpha("black", 0.3))

par(mfrow = c(4,1))
plot(density(sce_filt$Area))
plot(density(sqrt(sce_filt$Area)))
plot(density(log10(sce_filt$Area)))
plot(density(colSums(counts(sce_filt))))

plot(density(colSums(assay(sce_filt, "counts"))))
plot(density(colSums(assay(sce_filt, "logcounts_libsize"))))
plot(density(colSums(assay(sce_filt, "logcounts_area"))))
plot(density(colSums(assay(sce_filt, "logcounts_sqrtarea"))))

par(mfrow = c(2,2))
boxplot(colSums(assay(sce_filt, "counts")) ~ sce_filt$z + sce_filt$embryo,
        las = 2, xlab = "", main = "Counts")
boxplot(colSums(assay(sce_filt, "logcounts_libsize")) ~ sce_filt$z + sce_filt$embryo,
        las = 2, xlab = "", main = "Library size")
boxplot(colSums(assay(sce_filt, "logcounts_area")) ~ sce_filt$z + sce_filt$embryo,
        las = 2, xlab = "", main = "Area")
boxplot(colSums(assay(sce_filt, "logcounts_sqrtarea")) ~ sce_filt$z + sce_filt$embryo,
        las = 2, xlab = "", main = "sqrt Area")
```

There doesn't seem to be much of a difference, so we go with the most 
intuitive conceptually.

```{r}
assay(sce_filt, "logcounts") <- assay(sce_filt, "logcounts_area")

sce_filt
```

# PCA, batch correction, and UMAP

Now perform PCA, batch correction and UMAP calculation.

```{r}
assay(sce_filt, "cosineNorm") <- cosineNorm(assay(sce_filt, "logcounts"))

sce_filt <- scater::runPCA(sce_filt)

mbpca = multiBatchPCA(assay(sce_filt, "cosineNorm"),
                      batch = cData_filt$embryo_pos_z,
                      preserve.single = TRUE)
reducedDim(sce_filt, "mbPCA_cosine") <- mbpca[[1]]

out = reducedMNN(mbpca[[1]], batch = cData_filt$embryo_pos_z)

reducedDim(sce_filt, "mbPCA_cosine_MNN") <- out$corrected

sce_filt <- runUMAP(sce_filt, dimred = "mbPCA_cosine_MNN")
sce_filt$UMAP1 <- reducedDim(sce_filt, "UMAP")[,1]
sce_filt$UMAP2 <- reducedDim(sce_filt, "UMAP")[,2]

cData_filt$UMAP1 <- reducedDim(sce_filt, "UMAP")[,1]
cData_filt$UMAP2 <- reducedDim(sce_filt, "UMAP")[,2]
```

```{r}
ggplot(cData_filt,
       aes(x = UMAP1, y = UMAP2)) +
  theme_classic() +
  geom_point(alpha = 0.1, size = 1) +
  theme(legend.position = "bottom") +
  ggtitle("UMAP on all genes") +
  NULL

g = ggplot(cData_filt[sample(nrow(cData_filt)),],
           aes(x = UMAP1, y = UMAP2)) +
  theme_classic() +
  geom_point(aes(colour = embryo), alpha = 0.1, size = 1) +
  theme(legend.position = "bottom") +
  ggtitle("UMAP on all genes") +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 5))) +
  NULL
g

g + facet_wrap(~embryo)
```

```{r}
g = ggplot(cData_filt,
           aes(x = UMAP1, y = UMAP2)) +
  theme_classic() +
  geom_point(aes(colour = factor(z)),
             alpha = 0.1, size = 1) +
  theme(legend.position = "bottom") +
  ggtitle("UMAP on all genes") +
  scale_colour_discrete() +
  NULL
g

g + facet_wrap(~embryo)

ggplot(cData_filt,
       aes(x = UMAP1, y = UMAP2)) +
  theme_classic() +
  geom_point(aes(colour = log2(1+total)),
             size = 0.5) +
  theme(legend.position = "bottom") +
  scale_color_gradient2(low = "gray75", mid = "cornflowerblue", high = "black",
                        midpoint = 0.5*max(log2(1+cData_filt$total), na.rm = TRUE)) +
  ggtitle("UMAP on all genes") +
  NULL

ggplot(cData_filt,
       aes(x = UMAP1, y = UMAP2)) +
  theme_classic() +
  geom_point(aes(colour = detected),
             size = 0.5) +
  scale_color_gradient2(low = "gray75", mid = "cornflowerblue", high = "black",
                        midpoint = 0.5*max(cData_filt$detected, na.rm = TRUE)) +
  theme(legend.position = "bottom") +
  ggtitle("UMAP on all genes") +
  NULL
```

Plot PC values in space

```{r}
# plot PC values to space
for (pc_i in 1:20) {
  pc_vals = reducedDim(sce_filt, "mbPCA_cosine_MNN")[,pc_i]
  pc_cols = colorRampPalette(c("blue", "grey","red"))(100)[as.numeric(cut(pc_vals,100))]
  names(pc_cols) <- names(pc_vals)
  
  g = ggplot(data = boundary_polygons) + 
    geom_polygon(aes(x = segmentation_vertices_x_global_affine,
                     y = -segmentation_vertices_y_global_affine,
                     group = uniqueID,
                     fill = uniqueID),
                 colour = NA, size = 0.5) +
    facet_wrap(embryo~z, nrow = 2) +
    scale_fill_manual(values = pc_cols) +
    theme_classic() +
    coord_fixed() +
    theme(legend.position = "none") +
    ggtitle(paste0("PC: ", pc_i)) +
    NULL
  print(g)
}
```

# Save output

```{r}
sce_filt
dim(colData(sce_filt))
head(colData(sce_filt))
dim(cData_filt)
head(cData_filt)
dim(mRNA_df_filt)
head(mRNA_df_filt)
```

```{r, eval = FALSE}
saveRDS(sce_filt, file = "../analysis_output/E8.5/E8.5_sce_filt_unlabelled.Rds")
saveRDS(cData_filt, file = "../analysis_output/E8.5/E8.5_cData_filt_unlabelled.Rds")
saveRDS(mRNA_df_filt, file = "../analysis_output/E8.5/E8.5_mRNA_df_filt.Rds")
```

```{r}
sessionInfo()
```