---
title: "E8.5 seqFISH section experiment 120919"
author: "Shila Ghazanfar"
date: "29/05/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 10 
                     fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      cache = FALSE, cache.lazy = FALSE)
```

# Summary

This report generates processed data files for the SeqFISH experiment for 
embryo3. This is only for the seqFISH library (not smFISH rounds).

```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
```

```{r}
sapply(list.files("../functions/", full.names = TRUE), source)
```

# Set sample specific file paths - manual check

```{r}
fov_info_file = "../Data/TimEmbryos-120919/output/cellinfo/fovinfo.csv"
mRNA_files_dir = "../Data/TimEmbryos-120919/output/1error-sqrt6-pointsratio-2020-05-23/"
cell_files_dir = "../Data/TimEmbryos-120919/output/1error-sqrt6-pointsratio-2020-05-23/"
segmentation_files_dir = "../Data/TimEmbryos-120919/segmentation/"

processed_dir = "../Data/TimEmbryos-120919/processed/"
processed_suffix = "120919"

updateFiles = FALSE
```

# Load mRNA dot information

Convert the mRNA dot table into a SingleCellExperiment object. Using the cell
segmentation information.

```{r}
fov_info = read.csv(fov_info_file, header = TRUE, row.names = 1)
rownames(fov_info) <- paste0("fov.",rownames(fov_info))

mRNA_files = list.files(mRNA_files_dir, pattern = "segmented", full.names = TRUE)
mRNA_files
length(mRNA_files)

cell_files = list.files(cell_files_dir, pattern = "cellData", full.names = TRUE)
cell_files
length(cell_files)
```

# Set embryo and Estage labels - manual check

```{r}
embryo_label = rep("embryo3", length(cell_files))
names(embryo_label) <- paste0("fov.", 1:length(cell_files))

Estage_label = rep("E8.5", length(cell_files))
names(Estage_label) <- paste0("fov.", 1:length(cell_files))
```

# Get all genes, these are currently only seqFISH barcoded genes

```{r}
allgenes = sort(unique(unlist(sapply(mRNA_files, function(mRNA_file) {
  mRNA_df_raw = read.csv(mRNA_file, header = TRUE)
  return(as.character(unique(mRNA_df_raw$geneID)))
}, simplify = FALSE
))))
length(allgenes)
```

# Build params object

```{r}
params = list(
  fov_info_file = fov_info_file,
  mRNA_files_dir = mRNA_files_dir,
  cell_files_dir = cell_files_dir,
  segmentation_files_dir = segmentation_files_dir,
  processed_dir = processed_dir,
  processed_suffix = processed_suffix,
  fov_info = fov_info,
  mRNA_files = mRNA_files,
  cell_files = cell_files,
  embryo_label = embryo_label,
  Estage_label = Estage_label,
  allgenes = allgenes
)
```

# Function to process

Located in functions folder as `processSeqFISH`.

```{r}
out = processSeqFISH(params)
list2env(out, globalenv())
```

```{r}
sce = do.call(cbind, mRNA_sceList)
sce
head(colData(sce))
saveRDS(sce, file = paste0(
  processed_dir,
  "sce_all-",
  processed_suffix,
  ".Rds"))

mRNA_df = do.call(rbind, mRNA_dfList)
rownames(mRNA_df) <- NULL
dim(mRNA_df)
head(mRNA_df)
saveRDS(mRNA_df, file = paste0(
  processed_dir,
  "mRNA_df-",
  processed_suffix,
  ".Rds"))

boundary_polygons = getSegmentationVerticesDF(colData(sce))
dim(boundary_polygons)
head(boundary_polygons)
saveRDS(boundary_polygons, file = paste0(
  processed_dir,
  "boundary_polygons-",
  processed_suffix,
  ".Rds"))
```

```{r, fig.height = 24}
g = ggplot(boundary_polygons, aes(x = segmentation_vertices_x_global, 
                                  y = -segmentation_vertices_y_global)) + 
  geom_polygon(aes(group = uniqueID), show.legend = FALSE, 
               colour = "black", fill = NA, size = 0.2) + 
  theme_classic() +
  coord_fixed() +
  facet_wrap(~z)
print(g)

long_df = data.frame(fov = rep(rownames(fov_info), each = 4),
                     x = c(t(fov_info[,c("bound_x_1","bound_x_4", "bound_x_3", "bound_x_2")])),
                     y = c(t(fov_info[,c("bound_y_1","bound_y_4", "bound_y_3", "bound_y_2")]))
)

gg = g + 
  geom_polygon(aes(group = fov, x = x, y = -y), 
               fill = NA, colour = "red", size = 0.2,
               data = long_df, inherit.aes = FALSE) + 
  NULL

ggplot(long_df, aes(x = x, y = -y)) + 
  geom_polygon(fill = "black", aes(group = fov), colour = NA, size = 0.2) + 
  theme_classic() + 
  coord_fixed() + 
  NULL
print(gg)
```

```{r}
sessionInfo()
```