---
title: "E8.5 seqFISH section experiment 120919"
author: "Shila Ghazanfar"
date: "29/05/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 10 
                     fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      cache = FALSE, cache.lazy = FALSE)
```

# Summary

This report generates processed data files for the SeqFISH experiment for 
embryo3. This is only for the seqFISH library (not smFISH rounds), and includes 
the updated dot calls Nico provided on 25 May 2020.

```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
```

```{r}
source("spatial_functions.R")
source("segmentation_functions.R")
```

# Set sample specific file paths - manual check

```{r}
fov_info_file = "../Data/TimEmbryos-120919/output/cellinfo/fovinfo.csv"
mRNA_files_dir = "../Data/TimEmbryos-120919/output/1error-sqrt6-pointsratio-2020-05-23/"
cell_files_dir = "../Data/TimEmbryos-120919/output/1error-sqrt6-pointsratio-2020-05-23/"
segmentation_files_dir = "../Data/TimEmbryos-120919/segmentation/"

processed_dir = "../Data/TimEmbryos-120919/processed/"
processed_suffix = "120919"

updateFiles = FALSE
# updated_mRNA_files_dir = "../Data/TimEmbryos-120919/output/1error-sqrt6-lowerthreshold-2020-04-24/"
# updated_cell_files_dir = "../Data/TimEmbryos-120919/output/1error-sqrt6-lowerthreshold-2020-04-24/"
```

# Load mRNA dot information

Convert the mRNA dot table into a SingleCellExperiment object. Using the cell
segmentation information.

```{r}
fov_info = read.csv(fov_info_file, header = TRUE, row.names = 1)
rownames(fov_info) <- paste0("fov.",rownames(fov_info))

mRNA_files = list.files(mRNA_files_dir, pattern = "segmented", full.names = TRUE)
mRNA_files
length(mRNA_files)

cell_files = list.files(cell_files_dir, pattern = "cellData", full.names = TRUE)
cell_files
length(cell_files)
```

<!-- # Update mRNA files to new thresholding 24 April 2020 -->

```{r}
if (updateFiles) {
  updated_mRNA_files = list.files(updated_mRNA_files_dir,
                                  pattern = "segmented", full.names = TRUE)
  
  # order is very important!! first files are prioritised
  all_mRNA_files = c(updated_mRNA_files, mRNA_files)
  all_mRNA_files_pos = sapply(all_mRNA_files, function(x){
    pos_val_raw = unlist(strsplit(x, "-"))
    pos_val = grep("Pos[0-9]", pos_val_raw, value = TRUE)
    return(pos_val)
  })
  all_mRNA_files_pos
  
  # reassign the mRNA files
  mRNA_files <- all_mRNA_files[!duplicated(all_mRNA_files_pos)]
  mRNA_files
  length(mRNA_files)
}
```

<!-- # Update cell files to new thresholding 24 April 2020 -->

```{r}
if (updateFiles) {
  updated_cell_files = list.files(updated_cell_files_dir,
                                  pattern = "cellData", full.names = TRUE)
  
  # order is very important!! first files are prioritised
  all_cell_files = c(updated_cell_files, cell_files)
  all_cell_files_pos = sapply(all_cell_files, function(x){
    pos_val_raw = unlist(strsplit(x, "-"))
    pos_val = grep("Pos[0-9]", pos_val_raw, value = TRUE)
    return(pos_val)
  })
  all_cell_files_pos
  
  # reassign the cell files
  cell_files <- all_cell_files[!duplicated(all_cell_files_pos)]
  cell_files
  length(cell_files)
}
```

# Set embryo and Estage labels - manual check

```{r}
embryo_label = rep("embryo3", length(cell_files))
names(embryo_label) <- paste0("fov.", 1:length(cell_files))

Estage_label = rep("E8.5", length(cell_files))
names(Estage_label) <- paste0("fov.", 1:length(cell_files))
```

# Get all genes, these are currently only seqFISH barcoded genes

```{r}
allgenes = sort(unique(unlist(sapply(mRNA_files, function(mRNA_file) {
  mRNA_df_raw = read.csv(mRNA_file, header = TRUE)
  return(as.character(unique(mRNA_df_raw$geneID)))
}, simplify = FALSE
))))
length(allgenes)
```

# Build params object

```{r}
params = list(
  fov_info_file = fov_info_file,
  mRNA_files_dir = mRNA_files_dir,
  cell_files_dir = cell_files_dir,
  segmentation_files_dir = segmentation_files_dir,
  processed_dir = processed_dir,
  processed_suffix = processed_suffix,
  fov_info = fov_info,
  mRNA_files = mRNA_files,
  cell_files = cell_files,
  embryo_label = embryo_label,
  Estage_label = Estage_label,
  allgenes = allgenes
)
```

# Function to process

Located in spatial_functions.R as `processSeqFISH`.

```{r}
out = processSeqFISH(params)
list2env(out, globalenv())
```

```{r}
sce = do.call(cbind, mRNA_sceList)
sce
head(colData(sce))
saveRDS(sce, file = paste0(
  processed_dir,
  "sce_all-",
  processed_suffix,
  ".Rds"))

mRNA_df = do.call(rbind, mRNA_dfList)
rownames(mRNA_df) <- NULL
dim(mRNA_df)
head(mRNA_df)
saveRDS(mRNA_df, file = paste0(
  processed_dir,
  "mRNA_df-",
  processed_suffix,
  ".Rds"))

boundary_polygons = getSegmentationVerticesDF(colData(sce))
dim(boundary_polygons)
head(boundary_polygons)
saveRDS(boundary_polygons, file = paste0(
  processed_dir,
  "boundary_polygons-",
  processed_suffix,
  ".Rds"))
```

```{r, fig.height = 24}
g = ggplot(boundary_polygons, aes(x = segmentation_vertices_x_global, 
                                  y = -segmentation_vertices_y_global)) + 
  geom_polygon(aes(group = uniqueID), show.legend = FALSE, 
               colour = "black", fill = NA, size = 0.2) + 
  theme_classic() +
  coord_fixed() +
  facet_wrap(~z)
print(g)

long_df = data.frame(fov = rep(rownames(fov_info), each = 4),
                     x = c(t(fov_info[,c("bound_x_1","bound_x_4", "bound_x_3", "bound_x_2")])),
                     y = c(t(fov_info[,c("bound_y_1","bound_y_4", "bound_y_3", "bound_y_2")]))
)

gg = g + 
  geom_polygon(aes(group = fov, x = x, y = -y), 
               fill = NA, colour = "red", size = 0.2,
               data = long_df, inherit.aes = FALSE) + 
  NULL

ggplot(long_df, aes(x = x, y = -y)) + 
  geom_polygon(fill = "black", aes(group = fov), colour = NA, size = 0.2) + 
  theme_classic() + 
  coord_fixed() + 
  NULL
print(gg)
```

```{r}
sessionInfo()
```