---
title: "E8.5 seqFISH - Spatial neighbourhoods and cell compartments using polygon expansion"
author: "Shila Ghazanfar"
date: "11/06/2020"
output:
       html_document:
                     toc: true
                     toc_float:
                           collapsed: false
                           smooth_scroll: false
                     code_folding: hide
                     fig_width: 10 
                     fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE,cache.lazy = FALSE)
```

```{r}
set.seed(2020)
```

```{r}
library(SingleCellExperiment)
library(scater)
library(scran)
library(ComplexHeatmap)
library(ggplot2)
library(ggrepel)
library(GGally)
library(limma)
library(patchwork)
library(reshape)
library(gtools)
library(scattermore)
library(gganimate)
library(ggmap)
library(randomForest)
library(batchelor)
library(igraph)
library(cowplot)
library(ggpubr)
```

# TODO

# Summary

In this report we generate a network object that relates neighbours of cells,
by taking the cell segmentation polygon and expanding the vertices by a small 
expansion factor.

I also contract the cell segmentation 

Note that we're now only working with z-stacks 2 and 5. I place cells from these
two z-stacks onto the same plane and so also draw edges between them, these can 
be removed if we want by calling the induced subgraph per z-stack.

# Background code 

Load spatial and segmentation functions

```{r}
source("spatial_functions.R")
source("segmentation_functions.R")
source("celltype_colours.R")
source("neighbourSegments.R")
```

# Load data

```{r}
sce_filt = readRDS("../analysis_output/E8.5/E8.5_sce_filt_unlabelled.Rds")
# cData_filt = readRDS("../analysis_output/E8.5/E8.5_cData_filt.Rds")
cData_filt = as.data.frame(colData(sce_filt))
```

Set params

```{r}
expansionFactor = 1.3
contractionFactor = 0.75
```

<!-- Subset to z-stacks 2 and 5. -->

<!-- ```{r} -->
<!-- sce_filt <- sce[,sce$z %in% c(2,5)] -->
<!-- cData_filt <- cData[sce$z %in% c(2,5),] -->
<!-- ``` -->

Get polygons.

```{r}
boundary_polygons = getSegmentationVerticesDF(
    colData(sce_filt),
    xname = "segmentation_vertices_x_global_affine",
    yname = "segmentation_vertices_y_global_affine",
    othercols = c("uniqueID", "z", "embryo", "pos"))

dim(boundary_polygons)
head(boundary_polygons)
```

```{r}
g = ggplot(boundary_polygons,
           aes(x = segmentation_vertices_x_global_affine,
               y = -segmentation_vertices_y_global_affine,
               group = uniqueID)) + 
    facet_grid(z ~ embryo) + 
    theme_classic() +
    geom_polygon(alpha = 0.5, colour = "black", fill = "lightgrey") + 
    coord_fixed() + 
    NULL
g
```

# Illustrating polygon expansion using a small example

```{r}
cellIDs = c("embryo1_Pos0_cell100_z2",
            "embryo1_Pos0_cell101_z2",
            "embryo1_Pos0_cell130_z2")

boundary_polygons_example = subset(boundary_polygons, uniqueID %in% cellIDs)

length(unique(boundary_polygons_example$uniqueID))

# expansion
boundary_polygons_example <- addExpandedVertices(boundary_polygons_example,
                                                 expansionFactor = expansionFactor)

# contraction
boundary_polygons_example <- addExpandedVertices(boundary_polygons_example,
                                                 expansionFactor = contractionFactor,
                                                 new_xname = "segmentation_vertices_x_global_affine_contracted",
                                                 new_yname = "segmentation_vertices_y_global_affine_contracted")

df_point = as.data.frame(colData(sce_filt)[cellIDs,])
df_point$label = paste0("Cell ", 1:nrow(df_point))

g = ggplot(boundary_polygons_example,
           aes(x = segmentation_vertices_x_global_affine,
               y = -segmentation_vertices_y_global_affine,
               group = uniqueID)) + 
    # facet_grid(z ~ embryo) + 
    theme_classic() +
    geom_polygon(alpha = 0.5, colour = "black", fill = "lightgrey") + 
    coord_fixed() + 
    NULL
g

g1 = g + 
    geom_polygon(aes(x = segmentation_vertices_x_global_affine_expanded,
                     y = -segmentation_vertices_y_global_affine_expanded,
                     group = uniqueID),
                 fill = NA, colour = "blue",
                 inherit.aes = FALSE) +
    theme(axis.line = element_blank()) + 
    theme(axis.text = element_blank()) +
    theme(axis.ticks = element_blank()) +
    xlab("") + 
    ylab("") +
        geom_point(aes(x = x_global_affine, y = -y_global_affine), data = df_point, col = "darkgrey") +
    geom_text(aes(x = x_global_affine, y = -y_global_affine, label = label),
              data = df_point, size = 4, colour = "black") +
    NULL
g1

# g1 + 
    # geom_point(aes(x = x_global_affine, y = -y_global_affine), data = df_point) + 

# g = ggplot(df_point) +
#     # geom_point(aes(x = x_global_affine, y = -y_global_affine), data = df_point) + 
#     geom_text(aes(x = x_global_affine, y = -y_global_affine, label = label),
#               data = df_point, size = 10) +
#     theme_classic() + 
#     theme(axis.line = element_blank()) + 
#     theme(axis.text = element_blank()) +
#     theme(axis.ticks = element_blank()) +
#     xlab("") + 
#     ylab("") +
#     NULL
# g

ggsave(g1, file = "../Figures/cellneighbourhood_example.pdf", height = 4, width = 3)

# ggplot()

g1 = g + geom_polygon(aes(x = segmentation_vertices_x_global_affine_contracted,
                          y = -segmentation_vertices_y_global_affine_contracted,
                          group = uniqueID),
                      fill = NA, colour = "red",
                      inherit.aes = FALSE) +
    NULL
g1
```

Since the blue lines overlap with each other, these two cells will be marked 
as neighbouring, and an edge will be assigned to them in the output network.

The overlap function is also drawing edges between cells that are along the 
same z-axis. These can be pruned from the network later by taking the induced
subgraphs for each z-stack separately.

```{r}
# boundary_polygons_tmp = subset(boundary_polygons, pos %in% c("Pos0", "Pos50"))

# boundary_polygons_split = split.data.frame(boundary_polygons_tmp,
#                                            boundary_polygons_tmp$embryo)

# keeps z-stacks together
# boundary_polygons_split = split.data.frame(boundary_polygons,
#                                            boundary_polygons$embryo)

# keeps z-stacks separate
boundary_polygons_split = split.data.frame(boundary_polygons,
                                           interaction(boundary_polygons$embryo, 
                                                       boundary_polygons$z))

if (expansionFactor == 1) {
    boundary_polygons_split = split.data.frame(boundary_polygons,
                                               boundary_polygons$embryo)
}

verticesDFList = lapply(boundary_polygons_split, function(df) {
    split.data.frame(df,df$uniqueID)
})
```

# Calculate neighbourhood graph

Perform this calculation for each embryo separately. We can imagine performing
this calculation over multiple embryos with a large enough expansion factor, 
since the embryos are roughly aligned using affine transformation, so various 
tissue regions do overlap with each other.

```{r}
out_graph_filename = paste0("../analysis_output/E8.5/E8.5_neighbourGraph_",
                            expansionFactor,
                            ".Rds")

if (!file.exists(out_graph_filename)) {
    # function call for a single list element
    # out = neighbourVertices(verticesDFList[[1]],
    #                         xname = "segmentation_vertices_x_global_affine",
    #                         yname = "segmentation_vertices_y_global_affine",
    #                         expansionFactor = expansionFactor,
    #                         plot = FALSE,
    #                         plot2 = TRUE,
    #                         full = FALSE,
    #                         verbose = TRUE)
    
    if (FALSE) {
    out_list = lapply(verticesDFList, 
                      neighbourVertices,
                      xname = "segmentation_vertices_x_global_affine",
                      yname = "segmentation_vertices_y_global_affine",
                      expansionFactor = expansionFactor,
                      plot = FALSE,
                      plot2 = FALSE,
                      full = FALSE,
                      verbose = TRUE)
    } else {
     out_list = parallel::mclapply(verticesDFList, 
                      neighbourVertices,
                      xname = "segmentation_vertices_x_global_affine",
                      yname = "segmentation_vertices_y_global_affine",
                      expansionFactor = expansionFactor,
                      plot = FALSE,
                      plot2 = FALSE,
                      full = FALSE,
                      verbose = TRUE, mc.cores = 6)
    }
    
    out_graph = do.call(graph.union, out_list)
    saveRDS(out_graph, file = out_graph_filename)
}
```

# Classify mRNA dots from mRNA_df as inside or outside boundary regions

Load mRNA data. Subset down to z-stacks 2 and 5

```{r}
mRNA_df_filt = readRDS("../analysis_output/E8.5/E8.5_mRNA_df_filt.Rds")
# mRNA_df_filt <- subset(mRNA_df, z %in% c(2,5))
mRNA_df_filt$internal <- rep(FALSE, nrow(mRNA_df_filt))
```

Get contracted polygons, these are already subsetted to z-stacks 2 and 5.

```{r}
# contraction
boundary_polygons_contracted <- addExpandedVertices(boundary_polygons,
                                                 expansionFactor = contractionFactor,
                                                 new_xname = "segmentation_vertices_x_global_affine_contracted",
                                                 new_yname = "segmentation_vertices_y_global_affine_contracted")

verticesDFList_contracted = split.data.frame(boundary_polygons_contracted,
                                             boundary_polygons_contracted$uniqueID)
```

Take each polygon and class mRNA dots as to whether they are inside the polygon. 
This is an easier problem because we can automatically restrict to the mRNAs 
that are associated with the same cell. A value of TRUE means that the mRNA dot
is internal to the cell.

```{r}
mRNA_df_filt_filename = paste0("../analysis_output/E8.5/E8.5_mRNA_df_internal_",
                               contractionFactor,
                               ".Rds")

uniqueIDs = mixedsort(unique(as.character(boundary_polygons$uniqueID)))

if (!file.exists(mRNA_df_filt_filename)) {
    for (uniqueID in uniqueIDs) {
        
        print(uniqueID)
        
        mRNA_df_filt_ind = which(mRNA_df_filt$uniqueID == uniqueID)
        
        inPoly = inpolygon(
            mRNA_df_filt$x_global_affine[mRNA_df_filt_ind],
            mRNA_df_filt$y_global_affine[mRNA_df_filt_ind],
            verticesDFList_contracted[[uniqueID]][,"segmentation_vertices_x_global_affine_contracted"],
            verticesDFList_contracted[[uniqueID]][,"segmentation_vertices_y_global_affine_contracted"],
            boundary = TRUE)
        
        mRNA_df_filt[mRNA_df_filt_ind, "internal"] <- inPoly
        
    }
    
    saveRDS(mRNA_df_filt, file = mRNA_df_filt_filename)
} else {
    mRNA_df_filt <- readRDS(mRNA_df_filt_filename)
}
```

Examine this a little using the example from before

```{r}
table(mRNA_df_filt$internal)
head(mRNA_df_filt)

ggplot(subset(mRNA_df_filt, uniqueID %in% boundary_polygons_example$uniqueID),
       aes(x = x_global_affine, y = -y_global_affine,
           colour = internal)) +
    geom_point() + 
    geom_polygon(data = boundary_polygons_example,
                 aes(x = segmentation_vertices_x_global_affine,
                     y = -segmentation_vertices_y_global_affine,
                     group = uniqueID),
                 inherit.aes = FALSE,
                 colour = "black", fill = NA) +
    geom_polygon(data = boundary_polygons_example,
                 aes(x = segmentation_vertices_x_global_affine_contracted,
                     y = -segmentation_vertices_y_global_affine_contracted,
                     group = uniqueID),
                 inherit.aes = FALSE,
                 colour = "blue", fill = NA, linetype = "dotted") +
    theme_classic() +
    NULL
```

Some mRNA dots sit outside of the cell polygon boundary, but have been 
assigned to the cell - I think there's likely to be some extremely subtle 
shift in aligning the segmentation and coordinates, also that the segmentation 
is performed on a 512x512 grid so lacks some resolution.

Generate expression array with three dimensions, genes x cells x internal status

```{r}
counts_arr = tapply(mRNA_df_filt[,1], list(
    mRNA_df_filt$geneID,
    mRNA_df_filt$uniqueID,
    mRNA_df_filt$internal
), length)
counts_arr[is.na(counts_arr)] <- 0

class(counts_arr)
dim(counts_arr)
lapply(dimnames(counts_arr),head)
counts_arr[1:5,1:5,]
```

Brief comparison, add a jitter to be able to see points

```{r}
smoothScatter(c(jitter(counts_arr[,,"TRUE"], amount = 0.45)),
     c(jitter(counts_arr[,,"FALSE"], amount = 0.45)),
     pch = 16, cex = 0.5,
     col = scales::alpha("black", 0.2),
     xlab = "Internal", ylab = "Boundary", gridsize = 15)
abline(c(0,1))

plot(rowMeans(counts_arr[,,"TRUE"]),
     rowMeans(counts_arr[,,"FALSE"]), type = "n")
abline(c(0,1))
text(rowMeans(counts_arr[,,"TRUE"]),
     rowMeans(counts_arr[,,"FALSE"]),
     labels = rownames(counts_arr))
# names(which(rowMeans(counts_arr[,,"FALSE"]) > rowMeans(counts_arr[,,"TRUE"])))
```

Majority of genes have fewer counts in the interior than the boundaries, with
the exception of Meox1. Will be worth graphing MA-plots of the interior and 
boundary counts per mapped celltype, which I'll do in a separate report.

# Finish

```{r}
sessionInfo()
```